---
title: "Milan Urban Mobility Survey -- Cluster Analysis"
subtitle: "Unsupervised ML Analysis of Mixed-Type Survey Data"
author: "VL2 Research Team"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    theme: flatly
    fig-width: 10
    fig-height: 7
    df-print: paged
    self-contained: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

required_pkgs <- c(
  "tidyverse", "readr", "janitor", "scales", "viridis",
  "cluster", "FactoMineR", "factoextra", "clustMixType", "mclust",
  "corrplot", "patchwork", "ggrepel", "ggforce", "knitr",
  "tidygeocoder", "geosphere", "Rtsne", "fpc",
  "ggh4x", "ggtext"
)
missing <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(missing) > 0) install.packages(missing, repos = "https://cloud.r-project.org")

library(tidyverse)
library(cluster)
library(FactoMineR)
library(factoextra)
library(clustMixType)
library(mclust)
library(corrplot)
library(patchwork)
library(ggrepel)
library(ggforce)
library(scales)
library(viridis)
library(knitr)
library(tidygeocoder)
library(geosphere)
library(Rtsne)
library(fpc)
library(ggh4x)
library(ggtext)
theme_set(theme_minimal(base_size = 13))
set.seed(42)

# --- Global clustering parameter ---
# Researcher choice: k=6 for persona granularity. See "Choosing k" section.
n_clusters <- 6

# --- Global colour palette (qualitative, 6-safe) ---
# RColorBrewer "Dark2" is colour-blind friendly and distinct at 6+ levels.
cluster_palette <- RColorBrewer::brewer.pal(max(n_clusters, 3), "Dark2")
scale_color_clusters <- function(...) scale_color_manual(values = cluster_palette, ...)
scale_fill_clusters  <- function(...) scale_fill_manual(values = cluster_palette, ...)

```

# Introduction

This report presents an unsupervised clustering analysis of the VL2 Milan Urban Mobility Survey, conducted on March 14-15, 2026, with over 1,000 street interviews. The goal is to identify distinct citizen *personas* -- groups of respondents who share coherent patterns of attitudes, priorities, and behaviors regarding urban mobility, car usage, and public space in Milan.

The analysis proceeds in five stages. First, we **clean and preprocess** the raw survey data, handling the challenge of mixed variable types (opinion scales, ranked priorities, binary transport modes, categorical demographics, and spatial postal codes). Second, we **explore** the data distributions to understand what the sample looks like before any clustering. Third, we apply **four different clustering algorithms**, each suited to mixed-type data from a different angle, to ensure our findings are robust and not an artifact of a single method. Fourth, we **compare** the methods to assess how much they agree and which performs best. Finally, we **profile** the resulting clusters and translate them into interpretable personas that can inform urban policy discussions.

# Data Loading and Cleaning

The raw dataset contains 38 columns, of which we select 22 variables relevant to clustering. The remaining columns (timestamps, device info, audio transcriptions, interview coordinates) are metadata that do not describe the respondent's attitudes or characteristics.

## Load Raw Data

```{r}
#| label: load-data
raw <- read_csv("VL2 - Sheet1.csv", show_col_types = FALSE)
glimpse(raw)
```

## Select and Rename Variables

```{r}
#| label: select-rename
df <- raw |>
  select(
    spazio_auto      = `Troppo spazio per auto`,
    trasp_pub_ineff  = `Trasporto pubblico inefficiente`,
    meno_auto_clima  = `Meno auto per cambiamento climatico`,
    normale_auto     = `Normale auto per strada`,
    piu_verde        = `Più verde`,
    aggreg_disturbo  = `Aggregazione porta disturbo`,
    parcheggi_aggreg = `Da parcheggi ad aggregazione`,
    primo_posto      = `Primo posto`,
    secondo_posto    = `Secondo posto`,
    terzo_posto      = `Terzo posto`,
    limitare_auto    = `Limitare le auto migliora o peggiora`,
    lavoro           = Lavoro,
    genere           = Genere,
    auto             = Auto,
    bici             = Bici,
    mezzi_pub        = `Mezzi pubblici`,
    piedi            = Piedi,
    moto_scooter     = `Moto/scooter`,
    taxi             = Taxi,
    monopattino      = Monopattino,
    eta              = `Età`,
    cap              = CAP
  )
```

## Clean Age

The age field contains a few anomalous entries: one value of 20151 (a postal code entered in the wrong field) and a zero. These are set to NA.

```{r}
#| label: clean-age
df <- df |>
  mutate(
    eta = case_when(
      eta > 100 ~ NA_real_,
      eta == 0  ~ NA_real_,
      TRUE      ~ eta
    )
  )
```

## Clean CAP and Compute Distance to Milan Center

The CAP (Codice di Avviamento Postale) is the respondent's postal code of residence. As a raw number it is not directly usable for clustering -- "20121" is not meaningfully "closer" to "20122" than to "20143" in any metric that matters. To extract spatial information, we geocode each unique CAP via OpenStreetMap/Nominatim to obtain its geographic centroid, then compute the haversine (great-circle) distance to Piazza del Duomo (45.4642 N, 9.1900 E). This transforms a categorical postal code into a continuous variable -- *how far the respondent lives from Milan's center* -- which captures the urban-suburban gradient relevant to mobility attitudes. Distances greater than 100 km are treated as geocoding errors and set to NA.

```{r}
#| label: clean-cap-geocode
#| cache: true

# Step 1: Fix malformed CAPs
df <- df |>
  mutate(
    cap_clean = as.character(cap),
    cap_clean = str_pad(cap_clean, width = 5, side = "left", pad = "0"),
    cap_clean = if_else(nchar(cap_clean) > 5, NA_character_, cap_clean)
  )

# Step 2: Get unique valid CAPs
unique_caps <- df |>
  filter(!is.na(cap_clean)) |>
  distinct(cap_clean) |>
  rename(postal_code = cap_clean)

cat("Unique CAPs to geocode:", nrow(unique_caps), "\n")

# Step 3: Geocode via OSM/Nominatim (~1 req/sec, ~2 min total)
geocoded <- unique_caps |>
  mutate(query = paste(postal_code, "Italy", sep = ", ")) |>
  geocode(address = query, method = "osm", full_results = FALSE)

cat("Successfully geocoded:", sum(!is.na(geocoded$lat)), "of", nrow(geocoded), "CAPs\n")

# Step 4: Compute haversine distance to Piazza del Duomo
duomo <- c(lon = 9.1900, lat = 45.4642)

geocoded <- geocoded |>
  mutate(
    dist_duomo_km = if_else(
      !is.na(lat) & !is.na(long),
      as.numeric(distHaversine(
        cbind(long, lat),
        matrix(c(duomo["lon"], duomo["lat"]), nrow = n(), ncol = 2, byrow = TRUE)
      )) / 1000,
      NA_real_
    )
  )

# Step 5: Cap implausible distances (>100 km from Duomo is outside greater Lombardy)
n_outliers <- sum(geocoded$dist_duomo_km > 100, na.rm = TRUE)
geocoded <- geocoded |>
  mutate(dist_duomo_km = if_else(dist_duomo_km > 100, NA_real_, dist_duomo_km))

cat("Geocoding outliers (>100 km) set to NA:", n_outliers, "\n")

# Step 6: Join back to main data
df <- df |>
  left_join(
    geocoded |> select(postal_code, cap_lat = lat, cap_lon = long, dist_duomo_km),
    by = c("cap_clean" = "postal_code")
  )

# FALLBACK (uncomment if geocoding fails for many CAPs):
# Use interview GPS coordinates as spatial proxy
# df <- df |>
#   mutate(
#     dist_duomo_km = distHaversine(
#       cbind(raw$Longitude, raw$Latitude),
#       c(9.1900, 45.4642)
#     ) / 1000
#   )
```

## Missingness Report

Before clustering, we need to assess how much data is missing. Clustering algorithms generally cannot handle NA values, so we will use listwise deletion -- dropping any row with a missing value in any of the 22 clustering variables. This is acceptable when the complete-case rate is high (here ~88%), meaning we retain a large, representative sample.

```{r}
#| label: missingness
df |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing)) |>
  knitr::kable(caption = "Missing values per variable")
```

```{r}
#| label: listwise-deletion
clustering_vars <- c(
  "spazio_auto", "trasp_pub_ineff", "meno_auto_clima", "normale_auto",
  "piu_verde", "aggreg_disturbo", "parcheggi_aggreg",
  "primo_posto", "secondo_posto", "terzo_posto",
  "limitare_auto", "lavoro", "genere",
  "auto", "bici", "mezzi_pub", "piedi", "moto_scooter", "taxi", "monopattino",
  "eta", "dist_duomo_km"
)

df_complete <- df |> drop_na(all_of(clustering_vars))
cat("Rows retained for clustering:", nrow(df_complete), "of", nrow(df), "\n")
```

### Completeness Diagnostic

Listwise deletion assumes missingness is (approximately) completely at random (MCAR). If dropped respondents differ systematically from retained ones, the clustering sample is biased. The table below compares key summary statistics between retained and dropped cases.

```{r}
#| label: mcar-diagnostic
compare_vars <- c("spazio_auto", "trasp_pub_ineff", "meno_auto_clima",
                   "normale_auto", "piu_verde", "eta", "dist_duomo_km")

is_complete <- complete.cases(df |> select(all_of(clustering_vars)))

bind_rows(
  df |> filter(is_complete) |>
    summarise(group = "Retained", n = n(),
              across(all_of(compare_vars), ~ mean(., na.rm = TRUE))),
  df |> filter(!is_complete) |>
    summarise(group = "Dropped", n = n(),
              across(all_of(compare_vars), ~ mean(., na.rm = TRUE)))
) |>
  knitr::kable(digits = 2, caption = "Retained vs. Dropped Cases -- Mean Comparison")
```

# Variable Preprocessing

A key challenge in this analysis is that our variables are of fundamentally different types: Likert scales (1-10), binary indicators (0/1), unordered categories (employment, gender), and ranked text choices. Clustering algorithms need a consistent representation. This section transforms the raw variables into a format suitable for all four clustering methods. Note that the Likert scales (1-10) are treated as numeric throughout, which assumes equal spacing between scale points -- a standard but debatable assumption for ordinal data.

## Borda Scores for Ranked Priorities

The survey asked respondents to rank their top three priorities for the neighborhood from a list of six options (first, second, and third place). To convert these three text columns into a numeric representation that preserves ranking information, we use a **Borda count**: first place receives 3 points, second place 2 points, third place 1 point, and any category not ranked receives 0. This produces six numeric columns (one per priority category), each ranging from 0 to 3, that capture both *which* priorities a respondent cares about and *how much* they care (via the ranking).

```{r}
#| label: borda-scores
priority_categories <- c(
  "Migliorare la circolazione delle auto",
  "Mobilità sostenibile",
  "Più parcheggi",
  "Più verde urbano",
  "Spazi culturali e sport",
  "Spazi di socialità e famiglie"
)

priority_short <- c(
  "prio_circolazione", "prio_mobilita_sost", "prio_parcheggi",
  "prio_verde_urbano", "prio_cultura_sport", "prio_socialita"
)

for (i in seq_along(priority_categories)) {
  cat_name <- priority_categories[i]
  col_name <- priority_short[i]
  df_complete <- df_complete |>
    mutate(
      !!col_name := case_when(
        primo_posto   == cat_name ~ 3L,
        secondo_posto == cat_name ~ 2L,
        terzo_posto   == cat_name ~ 1L,
        TRUE                      ~ 0L
      )
    )
}
```

## Convert Categoricals to Factors

Categorical variables must be explicitly coded as R factors so that clustering algorithms treat them correctly (as categories, not numbers). We also merge the "Non binario" gender level (only 2 respondents) with "Preferisco non rispondere" to avoid instability from extremely rare categories.

```{r}
#| label: to-factors

# Merge rare genere level (Non binario, n=2) into Preferisco non rispondere
df_complete <- df_complete |>
  mutate(
    genere = if_else(genere == "Non binario", "Preferisco non rispondere", genere)
  )

df_complete <- df_complete |>
  mutate(
    limitare_auto = factor(limitare_auto),
    lavoro        = factor(lavoro),
    genere        = factor(genere),
    across(c(auto, bici, mezzi_pub, piedi, moto_scooter, taxi, monopattino),
           ~ factor(., levels = c(0, 1), labels = c("No", "Yes")))
  )
```

## Assemble Clustering Matrix

The final clustering matrix contains 25 columns: 15 numeric variables (7 Likert scales, age, distance from Duomo, and 6 Borda priority scores), 7 binary factors (transport modes), and 3 nominal factors (attitude toward car-limiting, employment, gender). This is the input to all four clustering methods.

```{r}
#| label: cluster-matrix
df_clust <- df_complete |>
  select(
    # Numeric / ordinal (treated as numeric)
    spazio_auto, trasp_pub_ineff, meno_auto_clima, normale_auto,
    piu_verde, aggreg_disturbo, parcheggi_aggreg,
    eta, dist_duomo_km,
    # Borda priority scores (numeric 0-3)
    all_of(priority_short),
    # Binary factors
    auto, bici, mezzi_pub, piedi, moto_scooter, taxi, monopattino,
    # Nominal factors
    limitare_auto, lavoro, genere
  )

cat("Clustering matrix dimensions:", nrow(df_clust), "x", ncol(df_clust), "\n")
str(df_clust)
```

# Exploratory Data Analysis

Before applying clustering algorithms, it is essential to understand the shape of the data. This section examines the distributions and correlations of all clustering variables. These plots help us anticipate what the algorithms will find: if certain variables are highly correlated, they will jointly drive cluster formation; if a variable has almost no variance, it will contribute little to distinguishing groups.

## Likert Scale Distributions

The seven Likert-scale variables (1-10) capture attitudes toward cars, public transport, green space, and social aggregation. The distributions reveal whether respondents tend to cluster at the extremes (polarized opinions) or spread across the scale.

```{r}
#| label: eda-likert
df_complete |>
  select(spazio_auto:parcheggi_aggreg) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_bar(fill = "steelblue", alpha = 0.8) +
  facet_wrap(~ variable, scales = "free_y", ncol = 2) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = "Distribution of Likert Scale Variables (1-10)",
       x = "Rating", y = "Count")
```

## Correlation Matrix

The correlation matrix shows how numeric variables relate to each other. Strong correlations (|r| > 0.3) indicate variables that move together and will jointly influence cluster membership. For instance, we might expect "too much space for cars" and "fewer cars for climate" to be positively correlated, as both reflect a pro-reduction attitude.

```{r}
#| label: eda-corr
#| fig-height: 9
df_complete |>
  select(spazio_auto:parcheggi_aggreg, eta, dist_duomo_km, all_of(priority_short)) |>
  cor(use = "pairwise.complete.obs") |>
  corrplot(method = "color", type = "lower", tl.cex = 0.7,
           addCoef.col = "black", number.cex = 0.6,
           title = "Correlation Matrix -- Numeric Variables",
           mar = c(0, 0, 2, 0))
```

## Categorical Variable Distributions

```{r}
#| label: eda-categorical
#| fig-height: 10
p1 <- df_complete |>
  dplyr::count(limitare_auto, name = "count") |>
  mutate(limitare_auto = fct_reorder(limitare_auto, count)) |>
  ggplot(aes(count, limitare_auto)) +
  geom_col(fill = "steelblue") +
  labs(title = "Limiting cars: improve or worsen?", x = "Count", y = NULL)

p2 <- df_complete |>
  dplyr::count(lavoro, name = "count") |>
  mutate(lavoro = fct_reorder(lavoro, count)) |>
  ggplot(aes(count, lavoro)) +
  geom_col(fill = "darkorange") +
  labs(title = "Employment", x = "Count", y = NULL)

p3 <- df_complete |>
  dplyr::count(genere, name = "count") |>
  mutate(genere = fct_reorder(genere, count)) |>
  ggplot(aes(count, genere)) +
  geom_col(fill = "forestgreen") +
  labs(title = "Gender", x = "Count", y = NULL)

p1 / p2 / p3
```

## Age and Distance Distributions

```{r}
#| label: eda-continuous
p_age <- ggplot(df_complete, aes(eta)) +
  geom_histogram(binwidth = 5, fill = "steelblue", alpha = 0.8) +
  labs(title = "Age Distribution", x = "Age", y = "Count")

# Reference lines for Milan's cerchie (approximate radii from Piazza del Duomo)
cerchie <- tribble(
  ~dist, ~label,
  1.2,   "Navigli",
  2.2,   "Bastioni",
  5.0,   "Filov. 90/91",
  7.0,   "Confine comunale"
)

p_dist <- ggplot(df_complete, aes(dist_duomo_km)) +
  geom_histogram(bins = 30, fill = "darkorange", alpha = 0.8) +
  geom_vline(data = cerchie, aes(xintercept = dist),
             linetype = "dashed", color = "grey30", linewidth = 0.4) +
  geom_text(data = cerchie, aes(x = dist, y = Inf, label = label),
            angle = 90, hjust = 1.1, vjust = -0.3, size = 2.8, color = "grey30") +
  labs(title = "Distance from Duomo (km) -- Full Range", x = "km", y = "Count")

p_dist_zoom <- ggplot(df_complete |> filter(dist_duomo_km <= 20), aes(dist_duomo_km)) +
  geom_histogram(binwidth = 1, fill = "darkorange", alpha = 0.8) +
  geom_vline(data = cerchie, aes(xintercept = dist),
             linetype = "dashed", color = "grey30", linewidth = 0.4) +
  geom_text(data = cerchie, aes(x = dist, y = Inf, label = label),
            angle = 90, hjust = 1.1, vjust = -0.3, size = 3, color = "grey30") +
  labs(title = "Distance from Duomo (km) -- Urban Core (< 20 km)", x = "km", y = "Count")

(p_age + p_dist) / p_dist_zoom
```

## Priority Borda Scores

This chart shows, on average, which priorities matter most to the sample. A mean Borda score of 2.0 would mean the category is typically ranked first or second; a score near 0 means it is rarely ranked at all.

```{r}
#| label: eda-priorities
df_complete |>
  select(all_of(priority_short)) |>
  pivot_longer(everything(), names_to = "category", values_to = "score") |>
  group_by(category) |>
  summarise(mean_score = mean(score), .groups = "drop") |>
  mutate(category = fct_reorder(category, mean_score)) |>
  ggplot(aes(mean_score, category)) +
  geom_col(fill = "mediumpurple") +
  labs(title = "Mean Borda Priority Score by Category",
       x = "Mean Score (0 = never ranked, 3 = first place)", y = NULL)
```

## Transport Mode Usage

Respondents could select multiple transport modes. This chart shows the prevalence of each mode in the sample. Modes with very low prevalence (taxi, monopattino) contribute minimal discriminative signal to clustering but are retained for completeness.

```{r}
#| label: eda-transport
df_complete |>
  select(auto, bici, mezzi_pub, piedi, moto_scooter, taxi, monopattino) |>
  pivot_longer(everything(), names_to = "mode", values_to = "used") |>
  group_by(mode) |>
  summarise(pct_yes = mean(used == "Yes"), .groups = "drop") |>
  mutate(mode = fct_reorder(mode, pct_yes)) |>
  ggplot(aes(pct_yes, mode)) +
  geom_col(fill = "coral") +
  scale_x_continuous(labels = percent_format()) +
  labs(title = "Transport Mode Usage (%)", x = "% of respondents", y = NULL)
```

# Clustering

## Why Four Methods?

No single clustering algorithm is universally best, especially with mixed-type data like ours (numeric scales, binary indicators, and unordered categories all in the same dataset). Different algorithms make different assumptions about what "similar" means and how clusters are shaped. By running four complementary methods and comparing their results, we can assess whether the personas we find are robust -- emerging consistently regardless of method -- or are artifacts of a particular algorithm's assumptions. If three out of four methods find roughly the same groups, we can be confident those groups are real.

## Choosing k

The number of clusters k is a researcher choice that balances statistical fit against narrative granularity. Purely statistical criteria (silhouette width) favour k = 2 for this data -- a blunt "pro-car vs. anti-car" split that is too coarse for policy-relevant personas. We set **k = `r n_clusters`** to allow fine-grained persona identification, and verify in the Robustness Check section below that this choice produces stable, interpretable clusters across multiple methods. The elbow, silhouette, gap, and BIC diagnostics for each method are shown for transparency, not as justification for our choice of k.

## Method 1: Gower Distance + PAM (k-Medoids)

**The idea**: The fundamental challenge with mixed data is defining "distance" between two respondents. How do you compare a Likert score of 7 to a gender of "Femminile"? The **Gower distance** solves this by computing a separate, type-appropriate distance for each variable (range-normalized difference for numeric, simple match/mismatch for categorical) and averaging them into a single dissimilarity score between 0 and 1. This produces a complete distance matrix between all pairs of respondents.

**PAM (Partitioning Around Medoids)** then clusters this distance matrix. Unlike k-means, which represents each cluster by a virtual mean point, PAM picks an actual respondent as the *medoid* (most central member) of each cluster. This is more robust to outliers and, importantly, the medoid is a real person whose survey responses we can inspect as the "prototype" of each cluster.

**Why this method matters**: Gower + PAM is the gold standard for mixed-type clustering. It requires no transformation of the data and makes minimal assumptions about cluster shapes. Its main limitation is computational: it computes an N x N distance matrix, which is feasible for our ~900 respondents but would not scale to millions.

### Compute Gower Distance

```{r}
#| label: gower-distance
#| cache: true
gower_dist <- daisy(df_clust, metric = "gower")
summary(gower_dist)
```

### Silhouette Analysis

The **silhouette width** measures how well each respondent fits their assigned cluster compared to the next-best alternative cluster. Values range from -1 (badly misclassified) to +1 (perfectly clustered). We compute the average silhouette for k = 2 through 8 to understand the trade-off between statistical separation and interpretive richness.

```{r}
#| label: gower-optimal-k
#| cache: true
sil_width_pam <- map_dbl(2:8, function(k) {
  pam_fit <- pam(gower_dist, k = k, diss = TRUE)
  pam_fit$silinfo$avg.width
})

tibble(k = 2:8, silhouette = sil_width_pam) |>
  ggplot(aes(k, silhouette)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 2:8) +
  labs(title = "PAM on Gower Distance -- Silhouette Width",
       x = "Number of Clusters (k)", y = "Average Silhouette Width")
```

### Fit PAM at k = `r n_clusters`

The silhouette plot above peaks at k = 2; we use k = `r n_clusters` for persona granularity (see "Choosing k" above).

```{r}
#| label: gower-pam-fit
# k fixed for cross-method comparison (see "Choosing k" section).
best_k_pam <- n_clusters
cat("Selected k for PAM:", best_k_pam, "\n")

pam_fit <- pam(gower_dist, k = best_k_pam, diss = TRUE)

df_complete <- df_complete |>
  mutate(cluster_pam = factor(pam_fit$clustering))
```

### Silhouette Plot

The per-observation silhouette plot shows how well each respondent fits their assigned cluster. Bars extending to the right indicate good fit; bars crossing zero mean the respondent is actually closer to another cluster's centroid than their own — by the silhouette criterion they are "misclassified", though this is common in survey data where attitudes form a continuum rather than discrete groups. The dashed horizontal line marks the overall average silhouette width; low values (near zero) indicate substantial overlap between clusters, reflecting gradual attitude transitions rather than sharp boundaries.

```{r}
#| label: gower-silhouette-plot
sil_obj <- silhouette(pam_fit$clustering, gower_dist)
fviz_silhouette(sil_obj) +
  labs(title = paste("PAM Silhouette Plot (k =", best_k_pam, ")"))
```

## Method 2: FAMD + k-Means

**The idea**: Instead of defining a custom distance (as Gower does), this approach first reduces the mixed-type data into a smaller set of continuous dimensions using **FAMD (Factor Analysis of Mixed Data)**. FAMD is a generalization of PCA that handles both numeric and categorical variables: it applies PCA-style decomposition to numeric variables and Multiple Correspondence Analysis (MCA) to categorical ones, balancing their contributions. The result is a set of continuous "factor scores" for each respondent -- new synthetic variables that capture the main axes of variation in the data.

Once the data is reduced to continuous scores, standard **k-means** clustering can be applied. K-means assigns each respondent to the cluster whose centroid (mean) is closest, iterating until assignments stabilize.

**Why this method matters**: FAMD + k-means complements Gower + PAM in two ways. First, the dimensionality reduction acts as a noise filter -- minor variables and random variation are absorbed into lower-ranked dimensions that we discard, so clustering focuses on the dominant patterns. Second, the FAMD variable maps provide a powerful interpretive tool: we can see *which variables drive the main axes of variation* and how clusters are positioned in this reduced space.

### Run FAMD

```{r}
#| label: famd-fit
#| cache: true
famd_result <- FAMD(df_clust, ncp = 10, graph = FALSE)

fviz_screeplot(famd_result, ncp = 10, addlabels = TRUE, barfill = "steelblue") +
  labs(title = "FAMD -- Screeplot (Variance Explained)")
```

### Variable Contributions

The variable maps show how each original variable relates to the FAMD dimensions. Variables pointing in the same direction are correlated; variables at right angles are independent. Dim 1--2 capture the strongest axis of variation (often age/employment vs. attitudes); Dim 3--4 reveal secondary structure that the first view misses. The per-dimension contribution bar charts below show which variables drive each dimension most strongly.

```{r}
#| label: famd-inspect
#| fig-height: 12
(fviz_famd_var(famd_result, repel = TRUE, col.var = "black") +
  labs(title = "FAMD -- Variable Map (Dim 1 vs Dim 2)")) /
(fviz_famd_var(famd_result, axes = c(3, 4), repel = TRUE, col.var = "black") +
  labs(title = "FAMD -- Variable Map (Dim 3 vs Dim 4)"))
```

```{r}
#| label: famd-contributions
#| fig-height: 10

# Helper: per-dimension contribution charts
make_quanti_contrib <- function(dim_idx) {
  tibble(
    variable = rownames(famd_result$quanti.var$contrib),
    contrib  = famd_result$quanti.var$contrib[, dim_idx]
  ) |>
    mutate(variable = fct_reorder(variable, contrib)) |>
    ggplot(aes(contrib, variable)) +
    geom_col(fill = "steelblue") +
    labs(title = paste("Quantitative -- Dim", dim_idx),
         x = "Contribution (%)", y = NULL)
}

make_quali_contrib <- function(dim_idx) {
  tibble(
    category = rownames(famd_result$quali.var$contrib),
    variable = rep(names(which(sapply(df_clust, is.factor))),
                   times = sapply(df_clust[, sapply(df_clust, is.factor)], nlevels)),
    contrib  = famd_result$quali.var$contrib[, dim_idx]
  ) |>
    mutate(label = paste0(variable, ": ", str_trunc(category, 35)),
           label = fct_reorder(label, contrib)) |>
    ggplot(aes(contrib, label)) +
    geom_col(fill = "darkorange") +
    labs(title = paste("Qualitative -- Dim", dim_idx),
         x = "Contribution (%)", y = NULL) +
    theme(axis.text.y = element_text(size = 7))
}

(make_quanti_contrib(1) + make_quanti_contrib(2)) /
(make_quanti_contrib(3) + make_quanti_contrib(4))
```

```{r}
#| label: famd-contributions-quali
#| fig-height: 14
#| fig-width: 14
(make_quali_contrib(1) + make_quali_contrib(2)) /
(make_quali_contrib(3) + make_quali_contrib(4))
```

### Optimal k Diagnostics

We show three standard diagnostics for choosing k in k-means: the **elbow plot** (total within-cluster sum of squares -- look for the "bend"), the **silhouette** (average cluster cohesion), and the **gap statistic** (comparison to a null reference distribution). These are shown for reference; we fix k = `r n_clusters` for comparability with the other methods.

The elbow plot shows a smoothly declining curve with no sharp "bend", indicating the data lacks a single dominant partition. The silhouette plot peaks at k = 2, reflecting a broad dichotomy (e.g., pro-car vs. anti-car attitudes). The gap statistic suggests a different optimal k. All three diagnostics favour fewer clusters than our chosen k = `r n_clusters`; we prefer the finer granularity because it yields policy-relevant sub-populations that a coarse split would obscure (see "Choosing k" above).

```{r}
#| label: famd-kmeans-optk
#| cache: true

# Retain enough dimensions for ~70% cumulative variance, or all available if threshold not reached
cumvar <- cumsum(famd_result$eig[, 2])
n_dims <- which(cumvar >= 70)[1]
if (is.na(n_dims)) n_dims <- nrow(famd_result$eig)
cat("Retaining", n_dims, "FAMD dimensions (", round(cumvar[n_dims], 1), "% variance)\n")

famd_scores <- as_tibble(famd_result$ind$coord[, 1:n_dims])

# Three diagnostic plots
p_elbow <- fviz_nbclust(famd_scores, kmeans, method = "wss", k.max = 8) +
  labs(title = "FAMD + k-Means -- Elbow (WSS)")
p_sil <- fviz_nbclust(famd_scores, kmeans, method = "silhouette", k.max = 8) +
  labs(title = "FAMD + k-Means -- Silhouette")
p_gap <- fviz_nbclust(famd_scores, kmeans, method = "gap_stat", k.max = 8,
                       nboot = 100) +
  labs(title = "FAMD + k-Means -- Gap Statistic")

p_elbow / p_sil / p_gap
```

```{r}
#| label: famd-select-k
# Fixed at n_clusters for cross-method comparability and persona interpretability
best_k_famd <- n_clusters
cat("Selected k for FAMD + k-Means:", best_k_famd, "\n")
```

### Fit k-Means on FAMD Scores

```{r}
#| label: famd-kmeans-fit
km_famd <- kmeans(famd_scores, centers = best_k_famd, nstart = 25, iter.max = 100)

df_complete <- df_complete |>
  mutate(cluster_famd = factor(km_famd$cluster))

fviz_cluster(km_famd, data = famd_scores,
             geom = "point", alpha = 0.5,
             palette = "Set2", ggtheme = theme_minimal()) +
  labs(title = paste("k-Means on FAMD Scores (k =", best_k_famd, ")"))
```

## Method 3: K-Prototypes

**The idea**: K-Prototypes is a direct extension of k-means designed for mixed data. For numeric variables it uses Euclidean distance (like k-means); for categorical variables it uses simple matching distance (counting mismatches). The two distances are combined with a weighting parameter that balances their contributions. Each cluster is represented by a *prototype*: a vector of means for numeric variables and modes for categorical ones.

**Why this method matters**: K-Prototypes works on the raw mixed data directly, without requiring either a distance matrix (like Gower) or dimensionality reduction (like FAMD). This makes it fast and conceptually simple. However, the trade-off is that Euclidean distance on raw numeric variables gives more weight to variables with larger scales. To avoid this, we z-score all numeric columns before running K-Prototypes, so that age and distance (ranges of ~70 and ~100) do not dominate Likert scales (range 9) and Borda scores (range 3).

### Prepare Standardized Data

```{r}
#| label: kproto-prep
# Z-score numeric columns so K-Prototypes treats all scales equally
num_cols <- names(df_clust)[sapply(df_clust, is.numeric)]
df_clust_std <- df_clust |>
  mutate(across(all_of(num_cols), ~ as.numeric(scale(.))))
str(df_clust_std)
```

### Elbow Diagnostic

```{r}
#| label: kproto-optk
#| cache: true
kproto_wss <- map_dbl(2:8, function(k) {
  kp <- kproto(df_clust_std, k = k, nstart = 5, verbose = FALSE)
  kp$tot.withinss
})

tibble(k = 2:8, wss = kproto_wss) |>
  ggplot(aes(k, wss)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 2:8) +
  labs(title = "K-Prototypes -- Elbow Plot (WSS)",
       x = "Number of Clusters (k)", y = "Total Within-Cluster Distance")
```

### Fit K-Prototypes at k = `r n_clusters`

```{r}
#| label: kproto-fit
# Fixed at n_clusters for cross-method comparability and persona interpretability
best_k_kproto <- n_clusters
cat("Selected k for K-Prototypes:", best_k_kproto, "\n")

kproto_fit <- kproto(df_clust_std, k = best_k_kproto, nstart = 10, verbose = FALSE)

df_complete <- df_complete |>
  mutate(cluster_kproto = factor(kproto_fit$cluster))

clprofiles(kproto_fit, df_clust_std)
```

## Method 4: Model-Based Clustering (Mclust) on FAMD Scores

**The idea**: Unlike the previous methods, which are heuristic (they optimize a distance-based criterion), **model-based clustering** assumes the data is generated by a mixture of Gaussian (bell-curve) distributions. Each cluster corresponds to one Gaussian component, characterized by its own mean, variance, and shape. The algorithm (Expectation-Maximization) finds the parameters that best explain the data, and the **BIC (Bayesian Information Criterion)** selects the best model geometry (spherical, ellipsoidal, etc.) and number of components.

Since Mclust requires continuous numeric input, we run it on the same FAMD factor scores used in Method 2. This means Mclust and FAMD + k-means operate on identical data, differing only in their clustering assumptions: k-means forces spherical clusters of equal size, while Mclust allows ellipsoidal clusters of varying size and orientation.

**Why this method matters**: Mclust provides a probabilistic perspective. Each respondent receives a probability of belonging to each cluster, not just a hard assignment. This is useful for identifying respondents who sit between personas. If Mclust's BIC-selected solution agrees with the other methods, it adds strong evidence that the cluster structure reflects real patterns rather than algorithmic assumptions.

Mclust's BIC provides an independent statistical estimate of k. We report it for transparency, then also fit at k = `r n_clusters` for the cross-method comparison.

```{r}
#| label: mclust-fit
#| cache: true
# Let BIC choose freely
mc_fit_bic <- Mclust(famd_scores, G = 2:8)
cat("BIC-optimal number of components:", mc_fit_bic$G, "\n")
cat("BIC-optimal model type:", mc_fit_bic$modelName, "\n")
summary(mc_fit_bic)
plot(mc_fit_bic, what = "BIC")

# Fit at n_clusters for cross-method comparison
mc_fit <- Mclust(famd_scores, G = n_clusters)
summary(mc_fit)

df_complete <- df_complete |>
  mutate(cluster_mclust = factor(mc_fit$classification))
```

Note: BIC selects G = `r mc_fit_bic$G` as optimal (`r mc_fit_bic$modelName` model), which may differ from our k = `r n_clusters` choice. See "Choosing k" above for rationale. We fit at k = `r n_clusters` for the cross-method comparison.

# Robustness Check: Cross-Method Agreement

We adopt **FAMD + k-Means** as our primary clustering solution: it operates on FAMD-reduced continuous scores (avoiding the scale sensitivity of K-Prototypes on raw data), and unlike PAM it is not evaluated on the same metric it was optimized on. The other three methods serve as robustness checks. If they find similar groups, the cluster structure reflects real patterns in the data; if they diverge, it may be method-dependent.

We use three complementary metrics. The **Adjusted Rand Index (ARI)** measures pairwise agreement between methods (1 = perfect agreement, 0 = random, negative = systematic disagreement). The **average silhouette width on Gower distance** provides a common quality benchmark (with caveats). Finally, **Jaccard bootstrap stability** assesses whether each cluster reappears when the data is resampled -- a direct test of robustness.

## Agreement Between Methods (Adjusted Rand Index)

```{r}
#| label: compare-agreement
methods <- list(
  PAM     = as.integer(df_complete$cluster_pam),
  FAMD_KM = as.integer(df_complete$cluster_famd),
  KProto  = as.integer(df_complete$cluster_kproto),
  MClust  = as.integer(df_complete$cluster_mclust)
)

ari_matrix <- matrix(NA, 4, 4, dimnames = list(names(methods), names(methods)))
for (i in 1:4) {
  for (j in 1:4) {
    ari_matrix[i, j] <- adjustedRandIndex(methods[[i]], methods[[j]])
  }
}

as_tibble(ari_matrix, rownames = "Method") |>
  knitr::kable(digits = 3, caption = "Adjusted Rand Index -- Pairwise Method Agreement")
```

## Silhouette Comparison on Gower Distance

We evaluate all four solutions using the same Gower distance matrix. **Caveat**: PAM was directly optimized on this Gower distance, so it has a structural advantage -- its silhouette score reflects optimization on the evaluation metric itself. The other methods were optimized on different representations (FAMD scores, raw mixed data). This comparison is informative but not a level playing field; the stability analysis below provides a fairer assessment.

```{r}
#| label: compare-silhouette
sil_compare <- tibble(
  Method = c("PAM", "FAMD+KMeans", "K-Prototypes", "Mclust"),
  Avg_Silhouette = c(
    mean(silhouette(as.integer(df_complete$cluster_pam),    gower_dist)[, 3]),
    mean(silhouette(as.integer(df_complete$cluster_famd),   gower_dist)[, 3]),
    mean(silhouette(as.integer(df_complete$cluster_kproto), gower_dist)[, 3]),
    mean(silhouette(as.integer(df_complete$cluster_mclust), gower_dist)[, 3])
  )
)

sil_compare |>
  mutate(Method = fct_reorder(Method, Avg_Silhouette)) |>
  ggplot(aes(Avg_Silhouette, Method)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(Avg_Silhouette, 3)), hjust = -0.1) +
  labs(title = "Average Silhouette Width (Gower Distance) by Method",
       subtitle = "Note: PAM is evaluated on the same metric it was optimized on",
       x = "Average Silhouette Width", y = NULL) +
  xlim(0, NA)

sil_compare |> knitr::kable(digits = 3)
```

## Cluster Stability (Jaccard Bootstrap)

The silhouette comparison above measures cluster *quality* but not *stability* -- whether the same clusters reappear under resampling. We use `fpc::clusterboot` with 100 bootstrap resamples to compute the mean Jaccard similarity for each cluster under each method. A Jaccard > 0.75 indicates a stable cluster; < 0.5 suggests the cluster is an artifact.

```{r}
#| label: compare-stability
#| cache: true

# PAM on Gower distance
cb_pam <- clusterboot(
  gower_dist, B = 100, bootmethod = "boot",
  clustermethod = claraCBI, usepam = TRUE, k = n_clusters,
  distances = TRUE, seed = 42, count = FALSE
)

# k-means on FAMD scores
cb_famd <- clusterboot(
  as.data.frame(famd_scores), B = 100, bootmethod = "boot",
  clustermethod = kmeansCBI, krange = n_clusters,
  seed = 42, count = FALSE
)

# Mclust on FAMD scores
cb_mclust <- clusterboot(
  as.data.frame(famd_scores), B = 100, bootmethod = "boot",
  clustermethod = noisemclustCBI, G = n_clusters,
  seed = 42, count = FALSE
)

stability_df <- tibble(
  Method = c("PAM", "FAMD+KMeans", "Mclust"),
  Mean_Jaccard = c(
    mean(cb_pam$bootmean),
    mean(cb_famd$bootmean),
    mean(cb_mclust$bootmean)
  ),
  Min_Jaccard = c(
    min(cb_pam$bootmean),
    min(cb_famd$bootmean),
    min(cb_mclust$bootmean)
  ),
  Cluster_Jaccards = c(
    paste(round(cb_pam$bootmean, 2), collapse = ", "),
    paste(round(cb_famd$bootmean, 2), collapse = ", "),
    paste(round(cb_mclust$bootmean, 2), collapse = ", ")
  )
)

stability_df |>
  knitr::kable(digits = 3,
               caption = "Jaccard Bootstrap Stability (100 resamples). >0.75 = stable, <0.50 = artifact.")

# Extract ARI and stability scalars for inline R in text below
ari_famd_pam    <- round(ari_matrix["FAMD_KM", "PAM"], 3)
ari_famd_mclust <- round(ari_matrix["FAMD_KM", "MClust"], 3)
ari_famd_kproto <- round(ari_matrix["FAMD_KM", "KProto"], 3)
famd_stability  <- round(mean(cb_famd$bootmean), 3)
```

Note: K-Prototypes is excluded from bootstrap stability because `clusterboot` does not natively support the `kproto` algorithm.

**Takeaway**: The ARI between FAMD+k-Means and the three comparison methods ranges from `r min(ari_famd_pam, ari_famd_mclust, ari_famd_kproto)` to `r max(ari_famd_pam, ari_famd_mclust, ari_famd_kproto)`, indicating `r if (max(ari_famd_pam, ari_famd_mclust, ari_famd_kproto) > 0.65) "substantial" else if (max(ari_famd_pam, ari_famd_mclust, ari_famd_kproto) > 0.4) "moderate" else "weak"` cross-method agreement. FAMD+k-Means achieves a mean Jaccard bootstrap stability of `r famd_stability`, `r if (famd_stability >= 0.75) "above" else "below"` the conventional 0.75 threshold for stable clusters. Overall, the cluster structure is `r if (famd_stability >= 0.75 && min(ari_famd_pam, ari_famd_mclust, ari_famd_kproto) > 0.4) "robust across methods and resampling" else if (famd_stability >= 0.75) "stable under resampling, though methods differ on some boundary respondents" else "partially method-dependent, with some clusters less stable than others"`.

## Primary Method Assignment

We use FAMD + k-Means as the primary clustering solution (see rationale above).

```{r}
#| label: select-best
df_complete <- df_complete |>
  mutate(cluster_final = cluster_famd)

# --- Data-driven cluster labeling (applied immediately for consistent names) ---
# Uses salience score = value² / sample_mean for each (cluster, feature) pair.
# This balances absolute prevalence (how common the behavior is in the cluster)
# with distinctiveness (how much it exceeds the sample average).
# Greedy assignment ensures each cluster gets a unique, descriptive label.

feature_labels <- c(
  pct_auto        = "Automobilisti",
  pct_mezzi       = "Utenti del Trasporto Pubblico",
  pct_bici        = "Ciclisti Urbani",
  pct_piedi       = "Pedoni di Quartiere",
  pct_moto        = "Motociclisti",
  spazio_auto     = "Critici dello Spazio Auto",
  meno_auto_clima = "Eco-sensibili",
  normale_auto    = "Difensori dell'Auto",
  trasp_pub_ineff = "Scettici del Trasporto Pubblico",
  piu_verde       = "Ambientalisti Urbani",
  pct_limit_yes   = "Pro Limitazione Auto"
)

label_profiles <- df_complete |>
  group_by(cluster_final) |>
  summarise(
    pct_auto        = mean(auto == "Yes"),
    pct_mezzi       = mean(mezzi_pub == "Yes"),
    pct_bici        = mean(bici == "Yes"),
    pct_piedi       = mean(piedi == "Yes"),
    pct_moto        = mean(moto_scooter == "Yes"),
    spazio_auto     = mean(spazio_auto) / 10,
    meno_auto_clima = mean(meno_auto_clima) / 10,
    normale_auto    = mean(normale_auto) / 10,
    trasp_pub_ineff = mean(trasp_pub_ineff) / 10,
    piu_verde       = mean(piu_verde) / 10,
    pct_limit_yes   = mean(limitare_auto == "Limiterei le auto per migliorare la qualità della vita in città"),
    .groups = "drop"
  )

sample_profile <- df_complete |>
  summarise(
    pct_auto        = mean(auto == "Yes"),
    pct_mezzi       = mean(mezzi_pub == "Yes"),
    pct_bici        = mean(bici == "Yes"),
    pct_piedi       = mean(piedi == "Yes"),
    pct_moto        = mean(moto_scooter == "Yes"),
    spazio_auto     = mean(spazio_auto) / 10,
    meno_auto_clima = mean(meno_auto_clima) / 10,
    normale_auto    = mean(normale_auto) / 10,
    trasp_pub_ineff = mean(trasp_pub_ineff) / 10,
    piu_verde       = mean(piu_verde) / 10,
    pct_limit_yes   = mean(limitare_auto == "Limiterei le auto per migliorare la qualità della vita in città")
  ) |>
  pivot_longer(everything(), names_to = "feature", values_to = "sample_mean")

profiles_long <- label_profiles |>
  pivot_longer(-cluster_final, names_to = "feature", values_to = "value") |>
  left_join(sample_profile, by = "feature") |>
  mutate(score = value^2 / pmax(sample_mean, 0.01))

cluster_name_vec <- setNames(character(nlevels(df_complete$cluster_final)),
                             levels(df_complete$cluster_final))
remaining_lbl <- profiles_long
for (i in seq_len(nlevels(df_complete$cluster_final))) {
  best <- remaining_lbl |> slice_max(score, n = 1, with_ties = FALSE)
  cluster_name_vec[as.character(best$cluster_final)] <- feature_labels[best$feature]
  remaining_lbl <- remaining_lbl |>
    filter(cluster_final != best$cluster_final, feature != best$feature)
}

# Relevel cluster_final with descriptive names
old_levels <- levels(df_complete$cluster_final)
df_complete <- df_complete |>
  mutate(cluster_final = factor(cluster_name_vec[as.character(cluster_final)],
                                levels = cluster_name_vec[old_levels]))

# Name the palette for consistent cluster↔colour mapping everywhere
names(cluster_palette) <- levels(df_complete$cluster_final)
scale_color_clusters <- function(...) scale_color_manual(values = cluster_palette, ...)
scale_fill_clusters  <- function(...) scale_fill_manual(values = cluster_palette, ...)

# Label lookup with sample sizes (for facet / legend labels)
cluster_sizes <- df_complete |> dplyr::count(cluster_final, name = "n")
label_lookup <- setNames(
  paste0(cluster_sizes$cluster_final, "\n(n = ", cluster_sizes$n, ")"),
  as.character(cluster_sizes$cluster_final)
)

# Display assignment for transparency
label_profiles |>
  mutate(Cluster = as.character(cluster_final),
         Label = cluster_name_vec[Cluster]) |>
  select(Cluster, Label, everything(), -cluster_final) |>
  knitr::kable(digits = 2,
               caption = "Cluster label assignment (salience = value\u00b2 / sample mean)")
```

## Export Clustered Data

```{r}
#| label: export-clustered-csv

# Identify which rows of df (= same order as raw) survived listwise deletion
complete_rows <- which(complete.cases(df[, clustering_vars]))

cluster_lookup <- tibble(
  raw_row      = complete_rows,
  cluster_id   = df_complete$cluster_famd,
  cluster_name = df_complete$cluster_final
)

raw_with_clusters <- raw |>
  mutate(raw_row = row_number()) |>
  left_join(cluster_lookup, by = "raw_row") |>
  select(-raw_row)

write_csv(raw_with_clusters, "VL2_clustered.csv")
cat("Exported", nrow(raw_with_clusters), "rows to VL2_clustered.csv\n")
cat("  With cluster:", sum(!is.na(raw_with_clusters$cluster_id)), "\n")
cat("  Without (dropped):", sum(is.na(raw_with_clusters$cluster_id)), "\n")
```

# Cluster Profiling

Clustering tells us *who goes with whom*, but not *why*. This section characterizes each cluster by examining how its members differ from the overall sample on every variable. We look at numeric means, categorical proportions, and demographic distributions to build a complete portrait of each persona.

## Numeric Variable Profiles

The table below reports the mean and standard deviation of each numeric variable within each cluster. Comparing these against the overall sample means (visible in the EDA section) reveals which attitudes and priorities distinguish each group.

```{r}
#| label: profile-numeric
numeric_vars <- c(
  "spazio_auto", "trasp_pub_ineff", "meno_auto_clima", "normale_auto",
  "piu_verde", "aggreg_disturbo", "parcheggi_aggreg",
  "eta", "dist_duomo_km",
  priority_short
)

profiles_num <- df_complete |>
  group_by(cluster_final) |>
  summarise(
    n = n(),
    across(all_of(numeric_vars), list(mean = mean, sd = sd), .names = "{.col}__{.fn}"),
    .groups = "drop"
  )

profiles_num |>
  knitr::kable(digits = 2, caption = "Cluster Profiles -- Numeric Variables (Mean +/- SD)")
```

## Heatmap of Standardized Cluster Means

The heatmap below standardizes each variable's cluster means into z-scores (number of standard deviations above or below the overall mean). This makes variables on different scales directly comparable. Red cells indicate the cluster scores notably above average; blue cells indicate below average. The raw mean values are printed inside each cell for reference.

```{r}
#| label: profile-heatmap
#| fig-height: 9
heatmap_vars <- c(
  "spazio_auto", "trasp_pub_ineff", "meno_auto_clima", "normale_auto",
  "piu_verde", "aggreg_disturbo", "parcheggi_aggreg",
  priority_short
)

df_complete |>
  group_by(cluster_final) |>
  summarise(across(all_of(heatmap_vars), mean), .groups = "drop") |>
  pivot_longer(-cluster_final, names_to = "variable", values_to = "mean_val") |>
  group_by(variable) |>
  mutate(z_score = (mean_val - mean(mean_val)) / sd(mean_val)) |>
  ungroup() |>
  ggplot(aes(cluster_final, variable, fill = z_score)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(mean_val, 1)), size = 3) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, name = "Z-score") +
  labs(title = "Cluster Profiles -- Standardized Mean Heatmap",
       x = "Cluster", y = NULL) +
  theme(axis.text.x = element_text(size = 8, angle = 25, hjust = 1),
        axis.text.y = element_text(size = 10))
```

## Categorical Variable Profiles

For categorical variables, we examine the proportion of each level within each cluster. Stacked bar charts make it easy to spot when a cluster has a markedly different composition -- for example, if one cluster is overwhelmingly composed of car users or students.

```{r}
#| label: profile-categorical
#| fig-height: 6
cat_vars <- c("limitare_auto", "lavoro", "genere",
              "auto", "bici", "mezzi_pub", "piedi", "moto_scooter")

for (v in cat_vars) {
  p <- df_complete |>
    dplyr::count(cluster_final, .data[[v]]) |>
    group_by(cluster_final) |>
    mutate(pct = n / sum(n)) |>
    ungroup() |>
    ggplot(aes(cluster_final, pct, fill = .data[[v]])) +
    geom_col(position = "fill") +
    scale_y_continuous(labels = percent_format()) +
    labs(title = paste("Cluster x", v), x = "Cluster", y = "Proportion") +
    theme(legend.position = "bottom",
          axis.text.x = element_text(size = 8, angle = 25, hjust = 1))
  print(p)
}
```

## Age and Distance Boxplots by Cluster

```{r}
#| label: profile-boxplots
p_age <- ggplot(df_complete, aes(cluster_final, eta, fill = cluster_final)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  scale_fill_clusters() +
  labs(title = "Age by Cluster", x = "Cluster", y = "Age") +
  theme(axis.text.x = element_text(size = 8, angle = 25, hjust = 1))

p_dist <- ggplot(df_complete, aes(cluster_final, dist_duomo_km, fill = cluster_final)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  scale_fill_clusters() +
  labs(title = "Distance from Duomo by Cluster", x = "Cluster", y = "km") +
  theme(axis.text.x = element_text(size = 8, angle = 25, hjust = 1))

p_age + p_dist
```

## Distance from Duomo by Cluster (Detailed)

The boxplot above compresses the distance distribution. This faceted histogram filters to the urban core (< 20 km) and overlays Milan's concentric cerchie (ring roads) as reference lines, matching the descriptive analysis above. This reveals whether different clusters draw from different parts of the city.

```{r}
#| label: profile-dist-duomo-cluster
#| fig-height: 8
#| fig-width: 10

cerchie_cluster <- tribble(
  ~dist, ~label,
  1.2,   "Navigli",
  2.2,   "Bastioni",
  5.0,   "Filov. 90/91",
  7.0,   "Confine comunale"
)

df_complete |>
  filter(dist_duomo_km <= 20) |>
  ggplot(aes(dist_duomo_km, fill = cluster_final, color = cluster_final)) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  geom_vline(data = cerchie_cluster, aes(xintercept = dist),
             linetype = "dashed", color = "grey30", linewidth = 0.4) +
  geom_text(data = cerchie_cluster, aes(x = dist, y = Inf, label = label),
            inherit.aes = FALSE,
            angle = 90, hjust = 1.1, vjust = -0.3, size = 2.8, color = "grey30") +
  scale_fill_clusters() +
  scale_color_clusters() +
  labs(title = "Distance from Duomo by Cluster (Urban Core < 20 km)",
       subtitle = "Dashed lines = Milan's concentric cerchie",
       x = "Distance from Duomo (km)", y = "Density",
       fill = "Cluster", color = "Cluster") +
  theme(legend.position = "bottom")
```

## Cluster Sizes

```{r}
#| label: profile-sizes
df_complete |>
  dplyr::count(cluster_final, name = "n") |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(cluster_final, n, fill = cluster_final)) +
  geom_col(show.legend = FALSE) +
  scale_fill_clusters() +
  geom_text(aes(label = paste0(n, " (", percent(pct, accuracy = 0.1), ")")),
            vjust = -0.3) +
  labs(title = "Cluster Sizes", x = "Cluster", y = "Count") +
  theme(axis.text.x = element_text(size = 9, angle = 25, hjust = 1))
```

# Cluster Comparison Visualizations

The profiling tables and stacked bars above provide detailed data, but can be hard to scan quickly. This section presents three summary visualizations designed for rapid comparison across clusters. The **diverging lollipop chart** highlights only each cluster's distinctive traits (what makes it different). The **radar chart** shows the full attitude profile shape for all clusters simultaneously. The **grouped bar charts** compare transport usage and car-limiting attitudes side by side.

```{r}
#| label: cluster-labels-reference
#| echo: false
# Labeling already applied in the select-best chunk above.
# label_lookup and cluster_palette are available for all downstream plots.
invisible(NULL)
```

## Diverging Lollipop: Distinctive Traits per Cluster

Each variable row shows how all clusters deviate from the sample average (only deviations with |z| > 0.3 shown). Clusters are distinguished by color; upward triangles with solid stems mark above-average values, downward triangles with dashed stems mark below-average values.

```{r}
#| label: lollipop-distinctive
#| fig-height: 10
#| fig-width: 8

# Compute cluster means for all numeric variables
numeric_vars_all <- c(
  "spazio_auto", "trasp_pub_ineff", "meno_auto_clima", "normale_auto",
  "piu_verde", "aggreg_disturbo", "parcheggi_aggreg", "eta", "dist_duomo_km",
  priority_short
)

# Build a unified numeric profile including binary/categorical as proportions
profile_data <- df_complete |>
  group_by(cluster_final) |>
  summarise(
    across(all_of(numeric_vars_all), mean, .names = "{.col}"),
    # Binary transport modes as % Yes
    `auto (% uso)` = mean(auto == "Yes") * 10,
    `bici (% uso)` = mean(bici == "Yes") * 10,
    `mezzi pubblici (% uso)` = mean(mezzi_pub == "Yes") * 10,
    `piedi (% uso)` = mean(piedi == "Yes") * 10,
    `moto/scooter (% uso)` = mean(moto_scooter == "Yes") * 10,
    # Key limitare_auto categories as proportion * 10 (for comparable scale)
    `limiterebbe senza condizioni (%)` = mean(limitare_auto == "Limiterei le auto per migliorare la qualità della vita in città") * 10,
    `auto necessarie (%)` = mean(limitare_auto == "Trovo che le auto siano necessarie per la vita quotidiana") * 10,
    .groups = "drop"
  )

# Overall sample means
sample_means <- df_complete |>
  summarise(
    across(all_of(numeric_vars_all), mean),
    `auto (% uso)` = mean(auto == "Yes") * 10,
    `bici (% uso)` = mean(bici == "Yes") * 10,
    `mezzi pubblici (% uso)` = mean(mezzi_pub == "Yes") * 10,
    `piedi (% uso)` = mean(piedi == "Yes") * 10,
    `moto/scooter (% uso)` = mean(moto_scooter == "Yes") * 10,
    `limiterebbe senza condizioni (%)` = mean(limitare_auto == "Limiterei le auto per migliorare la qualità della vita in città") * 10,
    `auto necessarie (%)` = mean(limitare_auto == "Trovo che le auto siano necessarie per la vita quotidiana") * 10
  )

sample_sds <- df_complete |>
  summarise(
    across(all_of(numeric_vars_all), sd),
    `auto (% uso)` = sd(as.numeric(auto == "Yes")) * 10,
    `bici (% uso)` = sd(as.numeric(bici == "Yes")) * 10,
    `mezzi pubblici (% uso)` = sd(as.numeric(mezzi_pub == "Yes")) * 10,
    `piedi (% uso)` = sd(as.numeric(piedi == "Yes")) * 10,
    `moto/scooter (% uso)` = sd(as.numeric(moto_scooter == "Yes")) * 10,
    `limiterebbe senza condizioni (%)` = sd(as.numeric(limitare_auto == "Limiterei le auto per migliorare la qualità della vita in città")) * 10,
    `auto necessarie (%)` = sd(as.numeric(limitare_auto == "Trovo che le auto siano necessarie per la vita quotidiana")) * 10
  )

# Compute z-scores for each cluster
all_vars <- names(sample_means)

# Variable-to-group mapping
var_group_lookup <- c(
  "spazio_auto" = "Atteggiamenti", "trasp_pub_ineff" = "Atteggiamenti",
  "meno_auto_clima" = "Atteggiamenti", "normale_auto" = "Atteggiamenti",
  "piu_verde" = "Atteggiamenti", "aggreg_disturbo" = "Atteggiamenti",
  "parcheggi_aggreg" = "Atteggiamenti",
  "eta" = "Demografia", "dist_duomo_km" = "Demografia",
  "prio_circolazione" = "Priorità", "prio_mobilita_sost" = "Priorità",
  "prio_parcheggi" = "Priorità", "prio_verde_urbano" = "Priorità",
  "prio_cultura_sport" = "Priorità", "prio_socialita" = "Priorità",
  "auto (% uso)" = "Trasporto", "bici (% uso)" = "Trasporto",
  "mezzi pubblici (% uso)" = "Trasporto", "piedi (% uso)" = "Trasporto",
  "moto/scooter (% uso)" = "Trasporto",
  "limiterebbe senza condizioni (%)" = "Politica",
  "auto necessarie (%)" = "Politica"
)

lollipop_df <- profile_data |>
  pivot_longer(-cluster_final, names_to = "variable", values_to = "cluster_mean") |>
  left_join(
    sample_means |> pivot_longer(everything(), names_to = "variable", values_to = "sample_mean"),
    by = "variable"
  ) |>
  left_join(
    sample_sds |> pivot_longer(everything(), names_to = "variable", values_to = "sample_sd"),
    by = "variable"
  ) |>
  mutate(
    z_score = (cluster_mean - sample_mean) / sample_sd,
    var_group = factor(var_group_lookup[variable],
                       levels = c("Trasporto", "Atteggiamenti", "Priorità",
                                  "Politica", "Demografia"))
  ) |>
  filter(abs(z_score) > 0.3) |>
  mutate(
    direction = if_else(z_score > 0, "Above average", "Below average")
  )

# Reorder variables by max |z| within each group
lollipop_df <- lollipop_df |>
  group_by(variable) |>
  mutate(max_abs_z = max(abs(z_score))) |>
  ungroup() |>
  mutate(variable = fct_reorder(variable, max_abs_z))

# Clean label_lookup for legend (replace newlines with spaces)
legend_labels <- setNames(gsub("\n", " ", label_lookup), names(label_lookup))

ggplot(lollipop_df, aes(x = z_score, y = variable,
                         color = cluster_final, fill = cluster_final,
                         group = cluster_final)) +
  geom_vline(xintercept = 0, linewidth = 0.5, color = "grey50") +
  geom_errorbarh(aes(xmin = 0, xmax = z_score, linetype = direction),
                 height = 0, linewidth = 1,
                 position = position_dodge(width = 0.7)) +
  geom_point(aes(shape = direction), size = 3,
             position = position_dodge(width = 0.7)) +
  scale_color_clusters(name = "Cluster", labels = legend_labels) +
  scale_fill_clusters(name = "Cluster", labels = legend_labels) +
  scale_shape_manual(name = "Deviation",
                     values = c("Above average" = 24, "Below average" = 25)) +
  scale_linetype_manual(name = "Deviation",
                        values = c("Above average" = "solid",
                                   "Below average" = "dotted")) +
  ggh4x::facet_grid2(var_group ~ ., scales = "free_y", space = "free_y",
                      switch = "y") +
  labs(
    title = "Distinctive Traits per Cluster (deviation from sample average)",
    subtitle = "Only variables with |z| > 0.3 shown. Rows grouped by variable type.",
    x = "Standard deviations from sample mean",
    y = NULL
  ) +
  theme(
    plot.title.position = "plot",
    legend.position = "top",
    legend.box = "horizontal",
    strip.text.y.left = element_text(size = 11, face = "bold", angle = 0, color = "black"),
    strip.background.y = element_rect(fill = "grey95", colour = "grey40"),
    strip.placement = "outside",
    axis.text.y = element_text(size = 10, color = "black"),
    axis.text.x = element_text(size = 10, color = "black"),
    panel.grid.major.y = element_blank(),
    panel.spacing.y = unit(0.5, "lines")
  ) +
  guides(
    color = guide_legend(order = 1, ncol = 1),
    fill = guide_legend(order = 1, ncol = 1),
    shape = guide_legend(order = 2),
    linetype = guide_legend(order = 2)
  )
```

## Radar Chart: Full Likert Profile Comparison

```{r}
#| label: radar-chart
#| fig-height: 12
#| fig-width: 12

# Radar chart -- one panel per cluster for clarity with 6 groups
likert_vars <- c("spazio_auto", "trasp_pub_ineff", "meno_auto_clima",
                 "normale_auto", "piu_verde", "aggreg_disturbo", "parcheggi_aggreg")

likert_labels <- c(
  "spazio_auto" = "Troppo spazio\nper auto",
  "trasp_pub_ineff" = "Trasp. pubblico\ninefficiente",
  "meno_auto_clima" = "Meno auto per\nil clima",
  "normale_auto" = "Normale avere\nauto",
  "piu_verde" = "Più verde",
  "aggreg_disturbo" = "Aggregazione\nporta disturbo",
  "parcheggi_aggreg" = "Parcheggi ad\naggregazione"
)

# Compute sample-wide mean as the grey reference polygon
sample_radar <- df_complete |>
  summarise(across(all_of(likert_vars), mean)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  mutate(var_id = match(variable, likert_vars))

radar_data <- df_complete |>
  group_by(cluster_final) |>
  summarise(across(all_of(likert_vars), mean), .groups = "drop") |>
  pivot_longer(all_of(likert_vars), names_to = "variable", values_to = "value") |>
  mutate(var_id = match(variable, likert_vars))

n_vars <- length(likert_vars)
angle_step <- 2 * pi / n_vars

add_radar_coords <- function(df) {
  df |> mutate(
    angle = (var_id - 1) * angle_step,
    x = value * cos(angle),
    y = value * sin(angle)
  )
}

radar_data   <- add_radar_coords(radar_data)
sample_radar <- add_radar_coords(sample_radar)

# Close polygons
close_poly <- function(df, grp) {
  df |> group_by(across(all_of(grp))) |>
    arrange(var_id) |>
    group_modify(~ bind_rows(.x, .x |> slice(1))) |>
    ungroup()
}

radar_poly  <- close_poly(radar_data, "cluster_final")
sample_poly <- bind_rows(sample_radar, sample_radar |> slice(1))

# Grid
grid_circles <- tibble(r = c(2, 4, 6, 8, 10)) |>
  mutate(data = purrr::map(r, ~ tibble(
    angle = seq(0, 2 * pi, length.out = 100),
    x = .x * cos(angle), y = .x * sin(angle)
  ))) |> unnest(data)

grid_spokes <- tibble(var_id = 1:n_vars) |>
  mutate(angle = (var_id - 1) * angle_step,
         xend = 10 * cos(angle), yend = 10 * sin(angle))

axis_labels <- tibble(
  var_id = 1:n_vars, variable = likert_vars,
  angle = (var_id - 1) * angle_step,
  x = 11.8 * cos(angle), y = 11.8 * sin(angle),
  label = likert_labels[variable]
)

ggplot() +
  geom_path(data = grid_circles, aes(x, y, group = r),
            color = "grey85", linewidth = 0.3) +
  geom_segment(data = grid_spokes, aes(x = 0, y = 0, xend = xend, yend = yend),
               color = "grey85", linewidth = 0.3) +
  # Sample average (grey reference)
  geom_polygon(data = sample_poly, aes(x, y),
               fill = "grey70", alpha = 0.15) +
  geom_path(data = sample_poly, aes(x, y),
            color = "grey50", linewidth = 0.5, linetype = "dashed") +
  # Cluster polygon
  geom_polygon(data = radar_poly, aes(x, y, fill = cluster_final),
               alpha = 0.25, show.legend = FALSE) +
  geom_path(data = radar_poly, aes(x, y, color = cluster_final),
            linewidth = 1, show.legend = FALSE) +
  geom_point(data = radar_data, aes(x, y, color = cluster_final),
             size = 2, show.legend = FALSE) +
  geom_text(data = axis_labels, aes(x, y, label = label),
            size = 2.5, lineheight = 0.85) +
  scale_color_clusters() +
  scale_fill_clusters() +
  coord_equal() +
  facet_wrap(~ cluster_final, ncol = 3,
            labeller = labeller(cluster_final = label_lookup)) +
  labs(title = "Likert Profile per Cluster (1-10 scale)",
       subtitle = "Grey dashed polygon = sample average") +
  theme_void(base_size = 11) +
  theme(
    strip.text = element_text(size = 9, face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10)
  )
```

## Grouped Bar Chart: Transport Mode Usage by Cluster

```{r}
#| label: transport-grouped-bars
#| fig-height: 6
#| fig-width: 10

transport_modes <- c("auto", "bici", "mezzi_pub", "piedi", "moto_scooter", "taxi", "monopattino")

transport_labels <- c(
  "auto" = "\U0001F697 Auto", "bici" = "\U0001F6B2 Bici",
  "mezzi_pub" = "\U0001F68C Mezzi pubblici", "piedi" = "\U0001F6B6 Piedi",
  "moto_scooter" = "\U0001F3CD Moto/scooter",
  "taxi" = "\U0001F695 Taxi", "monopattino" = "\U0001F6F4 Monopattino"
)

transport_data <- df_complete |>
  group_by(cluster_final) |>
  summarise(
    across(all_of(transport_modes), ~ mean(. == "Yes") * 100),
    n = n(),
    .groups = "drop"
  ) |>
  pivot_longer(all_of(transport_modes), names_to = "mode", values_to = "pct") |>
  mutate(mode_label = transport_labels[mode],
         mode_label = fct_reorder(mode_label, -pct))

# Sample average for reference
transport_avg <- df_complete |>
  summarise(across(all_of(transport_modes), ~ mean(. == "Yes") * 100)) |>
  pivot_longer(everything(), names_to = "mode", values_to = "pct_avg") |>
  mutate(mode_label = transport_labels[mode])

ggplot(transport_data, aes(x = mode_label, y = pct, fill = cluster_final)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7, alpha = 0.85) +
  geom_point(data = transport_avg, aes(x = mode_label, y = pct_avg),
             inherit.aes = FALSE, shape = 18, size = 4, color = "black") +
  scale_fill_clusters() +
  labs(
    title = "Transport Mode Usage by Cluster",
    subtitle = "Black diamonds = sample average",
    x = NULL, y = "% of cluster members", fill = "Cluster"
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "bottom",
    legend.text = element_text(size = 9)
  )
```

## Grouped Bar Chart: Attitude Toward Limiting Cars

```{r}
#| label: limitare-grouped-bars
#| fig-height: 7
#| fig-width: 11

# Shorter labels for the limitare_auto categories
limitare_short <- c(
  "Limiterei le auto per migliorare la qualità della vita in città" = "Limitare per migliorare\nqualità della vita",
  "Limiterei le auto SOLO se ci fossero buone alternative per muoversi" = "Limitare SOLO con\nbuone alternative",
  "Trovo che le auto siano necessarie per la vita quotidiana" = "Auto necessarie\nper vita quotidiana",
  "Non sento il bisogno di cambiare la situazione attuale" = "Non serve\ncambiare",
  "Non ho un'opinione precisa a riguardo" = "Nessuna opinione\nprecisa"
)

limitare_data <- df_complete |>
  dplyr::count(cluster_final, limitare_auto) |>
  group_by(cluster_final) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup() |>
  mutate(
    limitare_short = limitare_short[as.character(limitare_auto)],
    limitare_short = fct_reorder(limitare_short, -pct)
  )

ggplot(limitare_data, aes(x = limitare_short, y = pct, fill = cluster_final)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7, alpha = 0.85) +
  scale_fill_clusters() +
  labs(
    title = "Attitude Toward Limiting Cars by Cluster",
    x = NULL, y = "% of cluster members", fill = "Cluster"
  ) +
  theme(
    axis.text.x = element_text(size = 9),
    legend.position = "bottom",
    legend.text = element_text(size = 9)
  )
```

# Dimensionality Reduction Visualizations

The cluster profiles tell us *what* distinguishes each group, but not how clearly separated they are in the data. Dimensionality reduction visualizations project the high-dimensional respondent data into two dimensions for visual inspection. If clusters appear as distinct clouds, the separation is strong; if they overlap heavily, the clusters represent tendencies rather than sharp boundaries. Both outcomes are informative for understanding citizen attitudes.

## t-SNE on Gower Distance

**t-SNE (t-distributed Stochastic Neighbor Embedding)** is a nonlinear dimensionality reduction technique that excels at preserving local neighborhoods: respondents who are similar in the full 25-variable space will appear close together in the 2D plot. Unlike PCA, t-SNE can reveal non-convex cluster shapes. We run it directly on the Gower distance matrix.

```{r}
#| label: tsne-plot
#| cache: true
tsne_result <- Rtsne(gower_dist, is_distance = TRUE, perplexity = 30, dims = 2)

tsne_df <- tibble(
  tSNE1 = tsne_result$Y[, 1],
  tSNE2 = tsne_result$Y[, 2],
  Cluster = df_complete$cluster_final
)

ggplot(tsne_df, aes(tSNE1, tSNE2, color = Cluster, shape = Cluster)) +
  geom_point(alpha = 0.55, size = 2.5) +
  scale_color_clusters(name = "Cluster") +
  scale_shape_manual(name = "Cluster",
                     values = c(16, 17, 15, 18, 3, 4, 8, 1, 2)[seq_len(n_clusters)]) +
  labs(title = "t-SNE Visualization of Clusters (Gower Distance)",
       x = "t-SNE 1", y = "t-SNE 2") +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  theme(legend.position = "right",
        legend.text = element_text(size = 8))
```

## FAMD Biplot Colored by Cluster

The FAMD biplot shows respondents positioned on the first two factor dimensions (the same ones from Method 2). Unlike t-SNE, the axes are interpretable -- they correspond to the variable contributions shown earlier. The percentage of variance explained by each axis indicates how much of the total data variation is captured in this 2D view.

```{r}
#| label: famd-biplot
famd_df <- tibble(
  Dim1 = famd_result$ind$coord[, 1],
  Dim2 = famd_result$ind$coord[, 2],
  Cluster = df_complete$cluster_final
)

ggplot(famd_df, aes(Dim1, Dim2, color = Cluster, shape = Cluster)) +
  geom_point(alpha = 0.55, size = 2.5) +
  scale_color_clusters(name = "Cluster") +
  scale_shape_manual(name = "Cluster",
                     values = c(16, 17, 15, 18, 3, 4, 8, 1, 2)[seq_len(n_clusters)]) +
  labs(title = "FAMD Individual Map -- Colored by Final Cluster",
       x = paste0("Dim 1 (", round(famd_result$eig[1, 2], 1), "%)"),
       y = paste0("Dim 2 (", round(famd_result$eig[2, 2], 1), "%)")) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  theme(legend.text = element_text(size = 8))
```

## PAM Medoids (Representative Individuals)

The medoid is the most centrally located member of each PAM cluster -- the respondent whose overall profile is closest to all other members of their group. Examining medoids gives us a concrete, real-person embodiment of each persona.

```{r}
#| label: medoids
medoid_indices <- pam_fit$id.med
df_complete |>
  slice(medoid_indices) |>
  select(cluster_pam, all_of(names(df_clust))) |>
  knitr::kable(caption = "PAM Medoids -- Representative Individuals per Cluster")
```

# Cluster Interpretation

The `r n_clusters` clusters identified by the analysis represent distinct citizen archetypes with coherent attitudes toward urban mobility, environmental priorities, and car usage in Milan. The labels are assigned programmatically based on each cluster's most distinctive feature (see table above), so they remain correct across re-runs. Each persona profile below is generated dynamically from the data.

```{r}
#| label: interpretation-dynamic
#| results: asis

limit_solo <- "Limiterei le auto SOLO se ci fossero buone alternative per muoversi"
limit_yes  <- "Limiterei le auto per migliorare la qualità della vita in città"
limit_need <- "Trovo che le auto siano necessarie per la vita quotidiana"

transport_mode_labels <- c(
  auto = "car", bici = "bicycle", mezzi_pub = "public transit",
  piedi = "walking", moto_scooter = "moto/scooter"
)

for (cl in levels(df_complete$cluster_final)) {
  d <- df_complete |> filter(cluster_final == cl)
  cl_name <- cl
  n_cl <- nrow(d)
  pct_sample <- round(n_cl / nrow(df_complete) * 100, 1)

  # Dominant transport mode
  mode_pcts <- sapply(names(transport_mode_labels), function(m) mean(d[[m]] == "Yes"))
  top_mode <- names(which.max(mode_pcts))
  top_mode_pct <- round(max(mode_pcts) * 100)
  car_pct <- round(mean(d$auto == "Yes") * 100)

  # Attitudes
  m_spazio   <- round(mean(d$spazio_auto), 1)
  m_clima    <- round(mean(d$meno_auto_clima), 1)
  m_normale  <- round(mean(d$normale_auto), 1)
  m_trasp    <- round(mean(d$trasp_pub_ineff), 1)
  m_parcheg  <- round(mean(d$parcheggi_aggreg), 1)
  m_verde    <- round(mean(d$piu_verde), 1)

  # Car-limiting
  p_limit_yes  <- round(mean(d$limitare_auto == limit_yes) * 100)
  p_limit_solo <- round(mean(d$limitare_auto == limit_solo) * 100)
  p_limit_need <- round(mean(d$limitare_auto == limit_need) * 100)

  # Demographics
  m_eta <- round(mean(d$eta), 1)
  top_lavoro <- d |> dplyr::count(lavoro, sort = TRUE) |> slice(1)
  top_lav_pct <- round(top_lavoro$n / n_cl * 100)

  cat("\n## \"", cl_name, "\"\n\n", sep = "")

  cat("**", n_cl, " members (", pct_sample, "% of sample).** ",
      "Dominant transport: ", transport_mode_labels[top_mode],
      " (", top_mode_pct, "%)",
      if (top_mode != "auto") paste0("; car usage ", car_pct, "%") else "",
      ". Mean age: ", m_eta, ". ",
      "Largest employment group: ", as.character(top_lavoro$lavoro),
      " (", top_lav_pct, "%).\n\n", sep = "")

  cat("**Attitudes** (1-10 scale): ",
      "too much space for cars = ", m_spazio,
      ", fewer cars for climate = ", m_clima,
      ", normal to have cars = ", m_normale,
      ", public transport inefficient = ", m_trasp,
      ", more green space = ", m_verde,
      ", parking to social spaces = ", m_parcheg, ".\n\n", sep = "")

  cat("**Car-limiting policy**: ",
      p_limit_yes, "% would limit unconditionally, ",
      p_limit_solo, "% only with alternatives, ",
      p_limit_need, "% consider cars necessary.\n\n", sep = "")
}
```

# Personas: Individui Rappresentativi

I profili seguenti descrivono, per ciascun cluster, l'**individuo rappresentativo**: il rispondente reale che meglio incarna il proprio gruppo. La selezione avviene in due fasi: (1) si identificano i membri più centrali del cluster (top 10% per distanza di Gower complessiva più bassa), garantendo che il candidato sia geometricamente vicino a tutti gli altri; (2) tra questi, si sceglie quello che minimizza la discrepanza con il profilo aggregato del cluster (somma degli z-score in valore assoluto per le variabili numeriche + numero di scostamenti dalla moda per le variabili categoriali). Questo approccio produce un individuo che è sia centrale nel cluster sia fedele alle sue tendenze aggregate. Per ogni persona, si verifica la coerenza con le medie del cluster: eventuali scostamenti superiori a 1 deviazione standard (per le variabili numeriche) o rispetto alla moda (per le variabili categoriali) vengono segnalati.

## Calcolo dei medoidi e verifica di rappresentatività

```{r}
#| label: persona-medoids

# Convert Gower dissimilarity to full matrix (once)
gower_mat <- as.matrix(gower_dist)

# Variables used for the discrepancy score
numeric_vars_disc <- c(
  "spazio_auto", "trasp_pub_ineff", "meno_auto_clima", "normale_auto",
  "piu_verde", "aggreg_disturbo", "parcheggi_aggreg",
  "eta", "dist_duomo_km",
  priority_short
)
cat_vars_disc <- c("auto", "bici", "mezzi_pub", "piedi", "moto_scooter",
                    "taxi", "monopattino", "limitare_auto", "lavoro", "genere")

# Find representative individual for each FAMD+k-Means cluster:
#   1. Select the top 10% most central members (lowest total Gower distance)
#   2. Among those, pick the one minimizing discrepancy with cluster aggregates
#      (sum of |z-scores| for numeric vars + count of mode mismatches for categorical)
medoid_indices <- sapply(levels(df_complete$cluster_final), function(cl) {
  idx <- which(df_complete$cluster_final == cl)
  sub_mat <- gower_mat[idx, idx, drop = FALSE]
  total_dist <- rowSums(sub_mat)

  # Step 1: top 10% most central (at least 1 candidate)
  n_top <- max(1, ceiling(length(idx) * 0.10))
  top_local <- order(total_dist)[1:n_top]
  candidates <- idx[top_local]

  # Step 2: discrepancy score against cluster aggregates
  d <- df_complete[idx, ]

  # Pre-compute cluster means/SDs and modes
  cl_means <- sapply(numeric_vars_disc, function(v) mean(d[[v]], na.rm = TRUE))
  cl_sds   <- sapply(numeric_vars_disc, function(v) sd(d[[v]], na.rm = TRUE))
  cl_sds[is.na(cl_sds) | cl_sds == 0] <- Inf  # avoid division by zero; z = 0
  cl_modes <- sapply(cat_vars_disc, function(v) names(sort(table(d[[v]]), decreasing = TRUE))[1])

  discrepancy <- sapply(candidates, function(row_idx) {
    r <- df_complete[row_idx, ]
    # Numeric: sum of |z-scores|
    z_sum <- sum(abs((sapply(numeric_vars_disc, function(v) r[[v]]) - cl_means) / cl_sds))
    # Categorical: count of mode mismatches
    cat_miss <- sum(sapply(cat_vars_disc, function(v) as.character(r[[v]]) != cl_modes[v]))
    z_sum + cat_miss
  })

  candidates[which.min(discrepancy)]
})

# Extract representative rows
medoid_df <- df_complete |> slice(medoid_indices)

# --- Representativeness check ---
numeric_vars_check <- c(
  "spazio_auto", "trasp_pub_ineff", "meno_auto_clima", "normale_auto",
  "piu_verde", "aggreg_disturbo", "parcheggi_aggreg",
  "eta", "dist_duomo_km",
  priority_short
)

cat_vars_check <- c("auto", "bici", "mezzi_pub", "piedi", "moto_scooter",
                     "taxi", "monopattino", "limitare_auto", "lavoro", "genere")

repr_check <- map_dfr(levels(df_complete$cluster_final), function(cl) {
  d <- df_complete |> filter(cluster_final == cl)
  med_row <- medoid_df |> filter(cluster_final == cl)

  # Numeric variables
  num_rows <- map_dfr(numeric_vars_check, function(v) {
    cl_mean <- mean(d[[v]], na.rm = TRUE)
    cl_sd   <- sd(d[[v]], na.rm = TRUE)
    med_val <- med_row[[v]]
    z <- if (!is.na(cl_sd) && cl_sd > 0) (med_val - cl_mean) / cl_sd else 0
    tibble(
      cluster = cl, variable = v, type = "numeric",
      medoid_value = as.character(round(med_val, 1)),
      cluster_summary = paste0(round(cl_mean, 1), " (sd ", round(cl_sd, 1), ")"),
      z_score = round(z, 2),
      flag = if_else(abs(z) > 1, "deviazione", "OK")
    )
  })

  # Categorical variables
  cat_rows <- map_dfr(cat_vars_check, function(v) {
    mode_val <- names(sort(table(d[[v]]), decreasing = TRUE))[1]
    med_val  <- as.character(med_row[[v]])
    tibble(
      cluster = cl, variable = v, type = "categorical",
      medoid_value = med_val,
      cluster_summary = paste0("moda: ", mode_val),
      z_score = NA_real_,
      flag = if_else(med_val == mode_val, "OK", "deviazione")
    )
  })

  bind_rows(num_rows, cat_rows)
})
```

## Tabella di confronto: medoide vs. media del cluster

```{r}
#| label: persona-table

# Italian variable labels for display
var_labels_it <- c(
  eta = "Età", dist_duomo_km = "Distanza Duomo (km)",
  spazio_auto = "Troppo spazio auto", trasp_pub_ineff = "Trasp. pub. inefficiente",
  meno_auto_clima = "Meno auto per il clima", normale_auto = "Normale avere auto",
  piu_verde = "Più verde", aggreg_disturbo = "Aggregazione = disturbo",
  parcheggi_aggreg = "Parcheggi → socialità",
  prio_circolazione = "Prio: circolazione", prio_mobilita_sost = "Prio: mobilità sost.",
  prio_parcheggi = "Prio: parcheggi", prio_verde_urbano = "Prio: verde urbano",
  prio_cultura_sport = "Prio: cultura/sport", prio_socialita = "Prio: socialità",
  auto = "Usa auto", bici = "Usa bici", mezzi_pub = "Usa mezzi pub.",
  piedi = "A piedi", moto_scooter = "Moto/scooter",
  limitare_auto = "Limitare auto", lavoro = "Lavoro", genere = "Genere"
)

# Build comparison table
comparison_rows <- repr_check |>
  mutate(
    variable_it = if_else(variable %in% names(var_labels_it),
                          var_labels_it[variable], variable),
    confronto = case_when(
      flag == "deviazione" & type == "numeric" ~
        paste0("\u26a0 z = ", z_score),
      flag == "deviazione" & type == "categorical" ~
        "\u26a0 diverso dalla moda",
      TRUE ~ "\u2713"
    )
  ) |>
  select(Cluster = cluster, Variabile = variable_it,
         Medoide = medoid_value, `Media/Moda cluster` = cluster_summary,
         Verifica = confronto)

# Summary: deviations per cluster
dev_summary <- repr_check |>
  filter(flag == "deviazione") |>
  dplyr::count(cluster, name = "n_deviazioni")

dev_summary |>
  left_join(
    tibble(cluster = levels(df_complete$cluster_final),
           n_membri = as.integer(table(df_complete$cluster_final))),
    by = "cluster"
  ) |>
  knitr::kable(caption = "Numero di deviazioni per cluster",
               col.names = c("Cluster", "Deviazioni (su 25 variabili)", "Membri"))
```

::: {.callout-note collapse="true"}
## Dettaglio completo del confronto medoide vs. cluster

```{r}
#| label: persona-table-detail

comparison_rows |>
  knitr::kable(caption = "Confronto completo: individuo rappresentativo vs. tendenza centrale del cluster")
```
:::

## Profili narrativi per la comunicazione

I profili seguenti sono pensati come strumenti operativi per il team di comunicazione. Ogni persona sintetizza il proprio cluster in un ritratto narrativo che combina due livelli di informazione:

- **Dati individuali (medoide).** Il paragrafo di apertura — nome, età, professione, distanza dal Duomo, mezzi di trasporto utilizzati, posizione sulla limitazione delle auto — e le priorità per Milano descrivono l'**individuo rappresentativo** del cluster, selezionato con il metodo descritto sopra. Questo dà concretezza al profilo e lo rende più facilmente comunicabile.
- **Tratti distintivi del gruppo (medie del cluster).** La sezione «Cosa caratterizza questo gruppo» riporta solo le variabili per cui la media del cluster si discosta in modo rilevante dalla media campionaria (|z| > 0,3), le stesse evidenziate nel lollipop chart precedente. Questo garantisce che la caratterizzazione rifletta il comportamento collettivo del segmento, non le idiosincrasie di un singolo rispondente.

Per il dettaglio analitico completo (tutti i punteggi, scostamenti e z-score), si rimanda alla tabella di confronto nella sezione precedente.

```{r}
#| label: persona-narratives
#| results: asis
#| echo: false

# --- Name pools by gender ---
nomi_donna <- c("Giulia", "Francesca", "Chiara", "Alessandra", "Valentina", "Martina")
nomi_uomo  <- c("Marco", "Luca", "Alessandro", "Matteo", "Giovanni", "Andrea")
nomi_altro <- c("Alex", "Noa", "Sam", "Elia", "Gaia", "Mika")

set.seed(42)

# --- Italian transport mode labels ---
transport_it <- c(
  auto = "l'auto", bici = "la bicicletta", mezzi_pub = "i mezzi pubblici",
  piedi = "gli spostamenti a piedi", moto_scooter = "la moto/scooter",
  taxi = "il taxi", monopattino = "il monopattino elettrico"
)

# --- Likert variable descriptions (Italian, third person, no numeric scores) ---
likert_desc <- list(
  spazio_auto = list(
    low = "non ritiene che ci sia troppo spazio dedicato alle auto",
    mid = "è moderatamente critico/a verso lo spazio dedicato alle auto",
    high = "ritiene fortemente che ci sia troppo spazio per le auto in città"
  ),
  trasp_pub_ineff = list(
    low = "considera il trasporto pubblico sostanzialmente efficiente",
    mid = "ha un giudizio ambivalente sull'efficienza del trasporto pubblico",
    high = "considera il trasporto pubblico decisamente inefficiente"
  ),
  meno_auto_clima = list(
    low = "non è convinto/a che ridurre le auto aiuterebbe il clima",
    mid = "crede moderatamente che meno auto farebbero bene al clima",
    high = "crede fortemente che ridurre le auto sia importante per il clima"
  ),
  normale_auto = list(
    low = "non considera affatto normale avere auto per strada",
    mid = "ha un atteggiamento neutro verso la normalità delle auto in strada",
    high = "considera del tutto normale che ci siano auto in città"
  ),
  piu_verde = list(
    low = "non avverte una forte necessità di più verde urbano",
    mid = "desidera moderatamente più verde in città",
    high = "desidera fortemente più spazi verdi in città"
  ),
  aggreg_disturbo = list(
    low = "non associa i luoghi di aggregazione al disturbo",
    mid = "ha un atteggiamento ambivalente su aggregazione e disturbo",
    high = "associa fortemente i luoghi di aggregazione al disturbo"
  ),
  parcheggi_aggreg = list(
    low = "non è favorevole a convertire i parcheggi in spazi sociali",
    mid = "è moderatamente favorevole alla conversione dei parcheggi in spazi di socialità",
    high = "è molto favorevole a trasformare i parcheggi in spazi di aggregazione"
  )
)

likert_tier <- function(val) {
  if (val <= 3) "low" else if (val <= 6) "mid" else "high"
}

# --- Priority full names ---
prio_names_it <- setNames(priority_categories, priority_short)

# --- limitare_auto → natural language (third person) ---
limitare_frasi <- c(
  "Limiterei le auto per migliorare la qualità della vita in città" =
    "limiterebbe le auto senza condizioni per migliorare la qualità della vita in città",
  "Limiterei le auto SOLO se ci fossero buone alternative per muoversi" =
    "limiterebbe le auto solo a condizione che esistano buone alternative di trasporto",
  "Trovo che le auto siano necessarie per la vita quotidiana" =
    "ritiene che le auto siano necessarie per la vita quotidiana",
  "Non sento il bisogno di cambiare la situazione attuale" =
    "non sente il bisogno di cambiare la situazione attuale",
  "Non ho un'opinione precisa a riguardo" =
    "non ha un'opinione precisa a riguardo"
)

# --- Reuse lollipop_df for distinctive traits (all variable groups, |z|>0.3) ---
# lollipop_df was computed in the lollipop-distinctive chunk and contains z-scores
# for attitudes, transport %, priorities, policy, and demographics.

# --- Italian natural-language descriptions for ALL lollipop variables ---
# Maps each (variable, direction) pair to a third-person Italian phrase.
trait_desc_it <- list(
  # Attitudes (Likert)
  spazio_auto = list(
    above = "ritiene fortemente che ci sia troppo spazio per le auto in città",
    below = "non ritiene che ci sia troppo spazio dedicato alle auto"
  ),
  trasp_pub_ineff = list(
    above = "considera il trasporto pubblico decisamente inefficiente",
    below = "considera il trasporto pubblico sostanzialmente efficiente"
  ),
  meno_auto_clima = list(
    above = "crede fortemente che ridurre le auto sia importante per il clima",
    below = "non è convinto/a che ridurre le auto aiuterebbe il clima"
  ),
  normale_auto = list(
    above = "considera del tutto normale che ci siano auto in città",
    below = "non considera affatto normale avere auto per strada"
  ),
  piu_verde = list(
    above = "desidera fortemente più spazi verdi in città",
    below = "non avverte una forte necessità di più verde urbano"
  ),
  aggreg_disturbo = list(
    above = "associa fortemente i luoghi di aggregazione al disturbo",
    below = "non associa i luoghi di aggregazione al disturbo"
  ),
  parcheggi_aggreg = list(
    above = "è molto favorevole a trasformare i parcheggi in spazi di aggregazione",
    below = "non è favorevole a convertire i parcheggi in spazi sociali"
  ),
  # Transport modes
  `auto (% uso)` = list(
    above = "usa l'auto molto più della media",
    below = "usa l'auto molto meno della media"
  ),
  `bici (% uso)` = list(
    above = "usa la bicicletta molto più della media",
    below = "usa la bicicletta molto meno della media"
  ),
  `mezzi pubblici (% uso)` = list(
    above = "usa i mezzi pubblici molto più della media",
    below = "usa i mezzi pubblici molto meno della media"
  ),
  `piedi (% uso)` = list(
    above = "si sposta a piedi molto più della media",
    below = "si sposta a piedi molto meno della media"
  ),
  `moto/scooter (% uso)` = list(
    above = "usa la moto o lo scooter molto più della media",
    below = "usa la moto o lo scooter molto meno della media"
  ),
  # Priorities (Borda scores)
  prio_circolazione = list(
    above = "dà molta priorità al miglioramento della circolazione auto",
    below = "dà poca priorità al miglioramento della circolazione auto"
  ),
  prio_mobilita_sost = list(
    above = "dà molta priorità alla mobilità sostenibile",
    below = "dà poca priorità alla mobilità sostenibile"
  ),
  prio_parcheggi = list(
    above = "dà molta priorità a più parcheggi",
    below = "dà poca priorità a più parcheggi"
  ),
  prio_verde_urbano = list(
    above = "dà molta priorità al verde urbano",
    below = "dà poca priorità al verde urbano"
  ),
  prio_cultura_sport = list(
    above = "dà molta priorità a spazi culturali e sportivi",
    below = "dà poca priorità a spazi culturali e sportivi"
  ),
  prio_socialita = list(
    above = "dà molta priorità a spazi di socialità e famiglie",
    below = "dà poca priorità a spazi di socialità e famiglie"
  ),
  # Policy
  `limiterebbe senza condizioni (%)` = list(
    above = "è molto più propenso/a della media a limitare le auto senza condizioni",
    below = "è molto meno propenso/a della media a limitare le auto senza condizioni"
  ),
  `auto necessarie (%)` = list(
    above = "ritiene le auto necessarie per la vita quotidiana molto più della media",
    below = "ritiene le auto necessarie molto meno della media"
  ),
  # Demographics
  eta = list(
    above = "è mediamente più anziano/a della media del campione",
    below = "è mediamente più giovane della media del campione"
  ),
  dist_duomo_km = list(
    above = "vive mediamente più lontano dal centro rispetto al campione",
    below = "vive mediamente più vicino al centro rispetto al campione"
  )
)

# --- Generate one persona per cluster ---
for (i in seq_along(levels(df_complete$cluster_final))) {
  cl <- levels(df_complete$cluster_final)[i]
  med <- medoid_df |> filter(cluster_final == cl)
  n_cl <- sum(df_complete$cluster_final == cl)
  pct_cl <- round(n_cl / nrow(df_complete) * 100, 1)
  fraction_denom <- round(nrow(df_complete) / n_cl)

  # Assign name by gender
  gen <- as.character(med$genere)
  idx_name <- ((i - 1) %% 6) + 1
  nome <- if (gen == "Donna") nomi_donna[idx_name]
          else if (gen == "Uomo") nomi_uomo[idx_name]
          else nomi_altro[idx_name]
  suf <- if (gen == "Donna") "a" else if (gen == "Uomo") "o" else "\u0259"

  # --- Distinctive traits from lollipop_df (already filtered at |z|>0.3) ---
  cl_traits <- lollipop_df |>
    filter(cluster_final == cl) |>
    arrange(desc(abs(z_score)))

  # --- Transport modes ---
  modes_used <- names(transport_it)[sapply(names(transport_it), function(m) med[[m]] == "Yes")]
  modes_str <- if (length(modes_used) > 0) {
    paste(transport_it[modes_used], collapse = ", ")
  } else {
    "nessun mezzo abituale dichiarato"
  }

  # --- Car policy stance ---
  lim_val <- as.character(med$limitare_auto)
  lim_frase <- limitare_frasi[lim_val]
  if (is.na(lim_frase)) lim_frase <- paste0("ha risposto: \u00ab", lim_val, "\u00bb")

  # ===================== OUTPUT =====================

  # --- Header ---
  cat("\n### ", cl, "\n\n", sep = "")
  cat("*Circa 1 milanese su ", fraction_denom, " nel campione (n\u00a0=\u00a0",
      n_cl, ", ", pct_cl, "%)*\n\n", sep = "")

  # --- Flowing opening paragraph ---
  cat("**", nome, "** ha ", round(med$eta), " anni, è ",
      tolower(as.character(med$lavoro)),
      " e vive a circa ", round(med$dist_duomo_km, 1),
      " km dal Duomo. ", sep = "")
  cat("Per spostarsi in città utilizza ", modes_str, ". ", sep = "")
  cat("Riguardo alla politica sulle auto, ", nome, " ", lim_frase, ".\n\n", sep = "")

  # --- Distinctive traits (from lollipop, natural language) ---
  cat("**Cosa caratterizza questo gruppo.** Rispetto al campione complessivo, questo profilo si distingue perché ", sep = "")
  att_phrases <- character(0)
  for (j in seq_len(nrow(cl_traits))) {
    var <- as.character(cl_traits$variable[j])
    dir <- cl_traits$direction[j]
    dir_key <- if (dir == "Above average") "above" else "below"
    if (var %in% names(trait_desc_it) && dir_key %in% names(trait_desc_it[[var]])) {
      att_phrases <- c(att_phrases, trait_desc_it[[var]][[dir_key]])
    }
  }
  if (length(att_phrases) > 0) {
    cat(paste(att_phrases, collapse = "; "), ".\n\n", sep = "")
  } else {
    cat("non presenta tratti marcatamente diversi dalla media.\n\n")
  }

  # --- Priorities (from medoid) ---
  prio_vals <- sapply(priority_short, function(p) med[[p]])
  prio_order <- names(sort(prio_vals, decreasing = TRUE))
  top2 <- prio_order[1:min(2, length(prio_order))]
  if (prio_vals[top2[1]] == 0) {
    cat("**Le sue priorità per Milano.** Non ha indicato priorità particolari.\n\n")
  } else if (length(top2) == 2 && prio_vals[top2[2]] > 0) {
    cat("**Le sue priorità per Milano.** \u00ab", prio_names_it[top2[1]],
        "\u00bb e \u00ab", prio_names_it[top2[2]], "\u00bb.\n\n", sep = "")
  } else {
    cat("**Le sue priorità per Milano.** \u00ab", prio_names_it[top2[1]], "\u00bb.\n\n", sep = "")
  }

}
```

# Geographic Analysis: CAP-Level Variation

The cluster profiles above describe *who* the personas are, but not *where* they live. Distance from the Duomo is a useful proxy for urban-suburban gradients, yet it collapses Milan's rich spatial fabric into a single number. This section asks the complementary question: **do different neighborhoods (CAPs) host systematically different mixes of attitudes, behaviors, and priorities?**

We restrict the analysis to Milan-proper CAPs (201xx) with at least 15 respondents, providing a reasonable minimum for stable summary statistics. Milan CAPs below this threshold are pooled into a single "Altro Milano" row for context. Non-Milan and malformed CAPs are excluded.

```{r}
#| label: cap-analysis-prep

# --- Identify valid Milan CAPs with adequate sample size ---
cap_counts <- df_complete |>
  filter(grepl("^201", cap_clean)) |>
  dplyr::count(cap_clean, name = "n_cap")

valid_caps <- cap_counts |> filter(n_cap >= 15) |> pull(cap_clean)
small_caps <- cap_counts |> filter(n_cap < 15)  |> pull(cap_clean)

cat("Milan CAPs with n >= 15:", length(valid_caps), "\n")
cat("Milan CAPs with n <  15:", length(small_caps),
    "(total n =", sum(cap_counts$n_cap[cap_counts$cap_clean %in% small_caps]), ")\n")

# --- Subset and tag rows ---
df_cap <- df_complete |>
  filter(cap_clean %in% c(valid_caps, small_caps)) |>
  mutate(cap_group = if_else(cap_clean %in% valid_caps, cap_clean, "Altro Milano"))

# --- Order CAPs by distance from Duomo (center → periphery) ---
cap_order <- df_cap |>
  filter(cap_group != "Altro Milano") |>
  group_by(cap_group) |>
  summarise(mean_dist = mean(dist_duomo_km, na.rm = TRUE), .groups = "drop") |>
  arrange(mean_dist)

# "Altro Milano" goes at the bottom; "Campione" (overall) at the very top
cap_levels <- c("Campione", cap_order$cap_group, "Altro Milano")

# --- Build labels with sample sizes ---
cap_n <- df_cap |>
  dplyr::count(cap_group, name = "n_cap")

cap_label_lookup <- setNames(
  paste0(cap_n$cap_group, "  (n=", cap_n$n_cap, ")"),
  cap_n$cap_group
)
# Add the overall sample label
n_total_cap <- nrow(df_cap)
cap_label_lookup["Campione"] <- paste0("Campione  (n=", n_total_cap, ")")
```

## Cluster Composition by CAP

The stacked bar chart below shows the share of each persona within every CAP. CAPs are ordered from city center (top) to periphery (bottom). A "Campione" reference row shows the overall sample composition. Marked differences in the color mix across rows indicate that persona prevalence varies geographically.

```{r}
#| label: cap-cluster-composition
#| fig-height: 12
#| fig-width: 9

# Per-CAP cluster proportions
cap_cluster <- df_cap |>
  dplyr::count(cap_group, cluster_final) |>
  group_by(cap_group) |>
  mutate(pct = n / sum(n)) |>
  ungroup()

# Add "Campione" row (overall sample proportions)
sample_cluster <- df_cap |>
  dplyr::count(cluster_final) |>
  mutate(cap_group = "Campione", pct = n / sum(n))

cap_cluster_all <- bind_rows(cap_cluster, sample_cluster) |>
  mutate(
    cap_group = factor(cap_group, levels = cap_levels),
    cap_label = cap_label_lookup[as.character(cap_group)],
    cap_label = factor(cap_label, levels = cap_label_lookup[cap_levels])
  )

ggplot(cap_cluster_all, aes(pct, cap_label, fill = cluster_final)) +
  geom_col(position = "fill", width = 0.75) +
  geom_hline(
    yintercept = which(cap_levels == "Altro Milano") - 0.5,
    linetype = "dotted", color = "grey40"
  ) +
  geom_hline(
    yintercept = which(cap_levels == cap_order$cap_group[1]) - 0.5,
    linetype = "dotted", color = "grey40"
  ) +
  scale_x_continuous(labels = percent_format(), expand = expansion(mult = c(0, 0.02))) +
  scale_fill_clusters(name = "Cluster") +
  labs(
    title = "Composizione dei Cluster per CAP",
    subtitle = "CAP ordinati dal centro (alto) alla periferia (basso)",
    x = NULL, y = NULL
  ) +
  theme(
    axis.text.y = element_text(size = 8, family = "mono"),
    legend.position = "bottom",
    legend.text = element_text(size = 7),
    panel.grid.major.y = element_blank()
  ) +
  guides(fill = guide_legend(nrow = 2))
```

## Heatmap: Attitude and Behavior Profiles by CAP

The heatmap shows 12 selected dimensions (spanning attitudes, transport, priorities, policy, and demographics) across CAPs. Cell color encodes the z-score -- how many sample standard deviations the CAP mean deviates from the overall mean (red = above, blue = below). Raw mean values are printed inside each cell.

```{r}
#| label: cap-heatmap
#| fig-height: 14
#| fig-width: 12

# --- Compute per-CAP means for selected variables ---
cap_profiles <- df_cap |>
  group_by(cap_group) |>
  summarise(
    spazio_auto     = mean(spazio_auto),
    meno_auto_clima = mean(meno_auto_clima),
    normale_auto    = mean(normale_auto),
    trasp_pub_ineff = mean(trasp_pub_ineff),
    piu_verde       = mean(piu_verde),
    prio_mobilita_sost = mean(prio_mobilita_sost),
    prio_parcheggi  = mean(prio_parcheggi),
    `auto (%)`      = mean(auto == "Yes") * 100,
    `bici (%)`      = mean(bici == "Yes") * 100,
    `mezzi pub. (%)` = mean(mezzi_pub == "Yes") * 100,
    `pro limite auto (%)` = mean(limitare_auto == "Limiterei le auto per migliorare la qualità della vita in città") * 100,
    eta             = mean(eta),
    .groups = "drop"
  )

# Add "Campione" row
sample_row <- df_cap |>
  summarise(
    cap_group       = "Campione",
    spazio_auto     = mean(spazio_auto),
    meno_auto_clima = mean(meno_auto_clima),
    normale_auto    = mean(normale_auto),
    trasp_pub_ineff = mean(trasp_pub_ineff),
    piu_verde       = mean(piu_verde),
    prio_mobilita_sost = mean(prio_mobilita_sost),
    prio_parcheggi  = mean(prio_parcheggi),
    `auto (%)`      = mean(auto == "Yes") * 100,
    `bici (%)`      = mean(bici == "Yes") * 100,
    `mezzi pub. (%)` = mean(mezzi_pub == "Yes") * 100,
    `pro limite auto (%)` = mean(limitare_auto == "Limiterei le auto per migliorare la qualità della vita in città") * 100,
    eta             = mean(eta)
  )

cap_profiles_all <- bind_rows(cap_profiles, sample_row)

# --- Sample-level means and SDs for z-score computation ---
hm_vars <- setdiff(names(cap_profiles_all), "cap_group")

sample_means_cap <- df_cap |>
  summarise(
    spazio_auto     = mean(spazio_auto),
    meno_auto_clima = mean(meno_auto_clima),
    normale_auto    = mean(normale_auto),
    trasp_pub_ineff = mean(trasp_pub_ineff),
    piu_verde       = mean(piu_verde),
    prio_mobilita_sost = mean(prio_mobilita_sost),
    prio_parcheggi  = mean(prio_parcheggi),
    `auto (%)`      = mean(auto == "Yes") * 100,
    `bici (%)`      = mean(bici == "Yes") * 100,
    `mezzi pub. (%)` = mean(mezzi_pub == "Yes") * 100,
    `pro limite auto (%)` = mean(limitare_auto == "Limiterei le auto per migliorare la qualità della vita in città") * 100,
    eta             = mean(eta)
  ) |>
  pivot_longer(everything(), names_to = "variable", values_to = "sample_mean")

sample_sds_cap <- df_cap |>
  summarise(
    spazio_auto     = sd(spazio_auto),
    meno_auto_clima = sd(meno_auto_clima),
    normale_auto    = sd(normale_auto),
    trasp_pub_ineff = sd(trasp_pub_ineff),
    piu_verde       = sd(piu_verde),
    prio_mobilita_sost = sd(prio_mobilita_sost),
    prio_parcheggi  = sd(prio_parcheggi),
    `auto (%)`      = sd(as.numeric(auto == "Yes")) * 100,
    `bici (%)`      = sd(as.numeric(bici == "Yes")) * 100,
    `mezzi pub. (%)` = sd(as.numeric(mezzi_pub == "Yes")) * 100,
    `pro limite auto (%)` = sd(as.numeric(limitare_auto == "Limiterei le auto per migliorare la qualità della vita in città")) * 100,
    eta             = sd(eta)
  ) |>
  pivot_longer(everything(), names_to = "variable", values_to = "sample_sd")

# --- Variable grouping for x-axis facets ---
var_group_cap <- c(
  "spazio_auto" = "Atteggiamenti", "meno_auto_clima" = "Atteggiamenti",
  "normale_auto" = "Atteggiamenti", "trasp_pub_ineff" = "Atteggiamenti",
  "piu_verde" = "Atteggiamenti",
  "prio_mobilita_sost" = "Priorità", "prio_parcheggi" = "Priorità",
  "auto (%)" = "Trasporto", "bici (%)" = "Trasporto",
  "mezzi pub. (%)" = "Trasporto",
  "pro limite auto (%)" = "Politica",
  "eta" = "Demografia"
)

# --- Long format with z-scores ---
hm_long <- cap_profiles_all |>
  pivot_longer(-cap_group, names_to = "variable", values_to = "mean_val") |>
  left_join(sample_means_cap, by = "variable") |>
  left_join(sample_sds_cap, by = "variable") |>
  mutate(
    z_score = (mean_val - sample_mean) / sample_sd,
    cap_label = cap_label_lookup[cap_group],
    cap_label = factor(cap_label, levels = rev(cap_label_lookup[cap_levels])),
    var_group = factor(var_group_cap[variable],
                       levels = c("Atteggiamenti", "Priorità", "Trasporto",
                                  "Politica", "Demografia"))
  )

# --- Format display values ---
hm_long <- hm_long |>
  mutate(display_val = case_when(
    grepl("%", variable) ~ paste0(round(mean_val), "%"),
    variable == "eta"    ~ as.character(round(mean_val, 1)),
    TRUE                 ~ as.character(round(mean_val, 1))
  ))

ggplot(hm_long, aes(variable, cap_label, fill = z_score)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = display_val), size = 2.4) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0, name = "Z-score",
    limits = c(-2, 2), oob = scales::squish
  ) +
  facet_grid(. ~ var_group, scales = "free_x", space = "free_x") +
  labs(
    title = "Profilo per CAP -- Heatmap delle Deviazioni dalla Media Campionaria",
    subtitle = "Rosso = sopra la media, Blu = sotto la media. Valori grezzi nelle celle.",
    x = NULL, y = NULL
  ) +
  theme(
    axis.text.x.top = element_text(size = 7.5, angle = 40, hjust = 0, margin = margin(b = 5)),
    axis.text.y = element_text(size = 7, family = "mono"),
    strip.text  = element_text(size = 9, face = "bold", margin = margin(b = 8)),
    strip.placement = "outside",
    panel.grid  = element_blank(),
    legend.position = "right"
  ) +
  scale_x_discrete(position = "top")
```

## Spatial Map: Cluster Composition by CAP

Each pie chart represents a CAP centroid at its geocoded location, showing the full cluster composition (not just the dominant one). Pie size is proportional to sample size. Dashed circles mark Milan's cerchie (ring roads) for geographic reference.

```{r}
#| label: cap-spatial-map
#| fig-height: 10
#| fig-width: 10

library(scatterpie)

# --- Cluster proportions and coordinates per CAP ---
cap_geo <- df_cap |>
  filter(cap_group != "Altro Milano") |>
  group_by(cap_group) |>
  summarise(
    lon = mean(cap_lon, na.rm = TRUE),
    lat = mean(cap_lat, na.rm = TRUE),
    n_cap = n(),
    .groups = "drop"
  )

# Wide-format cluster proportions for scatterpie
cluster_lvls <- levels(df_complete$cluster_final)

cap_wide <- df_cap |>
  filter(cap_group != "Altro Milano") |>
  dplyr::count(cap_group, cluster_final) |>
  group_by(cap_group) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  select(-n) |>
  pivot_wider(names_from = cluster_final, values_from = pct, values_fill = 0) |>
  left_join(cap_geo, by = "cap_group")

# Pie radius proportional to sqrt(n) for area-proportional scaling
cap_wide <- cap_wide |>
  mutate(radius = 0.002 + 0.003 * sqrt(n_cap) / max(sqrt(n_cap)))

# --- Cerchie as circles around Duomo ---
duomo_lon <- 9.1900
duomo_lat <- 45.4642

km_to_lat <- 1 / 111
km_to_lon <- 1 / (111 * cos(duomo_lat * pi / 180))

cerchie_circles <- cerchie |>
  crossing(theta = seq(0, 2 * pi, length.out = 100)) |>
  mutate(
    lon = duomo_lon + dist * km_to_lon * cos(theta),
    lat = duomo_lat + dist * km_to_lat * sin(theta)
  )

ggplot() +
  # Cerchie reference circles
  geom_path(
    data = cerchie_circles,
    aes(lon, lat, group = label),
    linetype = "dashed", color = "grey60", linewidth = 0.3
  ) +
  geom_text(
    data = cerchie |> mutate(
      lon = duomo_lon + dist * km_to_lon,
      lat = duomo_lat + 0.002
    ),
    aes(lon, lat, label = label),
    size = 2.5, color = "grey40", hjust = 0, nudge_x = 0.002
  ) +
  # Duomo center marker
  annotate("point", x = duomo_lon, y = duomo_lat,
           shape = 4, size = 3, stroke = 1, color = "black") +
  # Pie charts at each CAP centroid
  geom_scatterpie(
    data = cap_wide,
    aes(x = lon, y = lat, r = radius),
    cols = cluster_lvls,
    color = "grey30", linewidth = 0.2
  ) +
  # CAP labels
  geom_text(
    data = cap_wide,
    aes(lon, lat, label = cap_group),
    size = 1.8, nudge_y = 0.008, color = "grey20"
  ) +
  scale_fill_clusters(name = "Cluster") +
  coord_fixed(ratio = 1 / cos(duomo_lat * pi / 180)) +
  labs(
    title = "Mappa dei CAP -- Composizione dei Cluster per Zona",
    subtitle = "Dimensione torta = n rispondenti; cerchi tratteggiati = cerchie di Milano; X = Piazza del Duomo"
  ) +
  theme_void(base_size = 12) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9, color = "grey40")
  )
```
